---
title: "Dundee_finalAnalysis"
output: html_document
date: "2022-11-18"
editor_options: 
  chunk_output_type: console
---

```{r}
# clean global R environment
rm(list = ls())

options(scipen = 9999)

# required R packages

library(readxl)
library(dplyr)
library(rstatix)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(readr)
library(funrar)
library(vegan)
library(ggrepel)
library(gplots)
library(treemap)
library(compositions)
library(cowplot)
library(pheatmap)


library(forcats)
library(factoextra)
library(ggridges)
```

#load files
```{r}
metadata <- read_excel("working_metadataDundee.xlsx")

metadata <- data.frame(metadata)
rownames(metadata) <- metadata$Lab_ID

#groups based on Chalmers 2013, The Bronchiectasis Severity Index

metadata$age_group <- with(metadata,
                           ifelse(metadata$age < 50, "A",
                           ifelse(metadata$age >= 50 & metadata$age < 70, "B",
                           ifelse(metadata$age >= 70 & metadata$age < 80, "C",
                           ifelse(metadata$age >= 80, "D", NA)))))

metadata$BMI_group <- with(metadata,
                     ifelse(metadata$BMI < 18.5, "A",
                     ifelse(metadata$BMI <= 25 & metadata$BMI >= 18.5, "B",
                     ifelse(metadata$BMI <= 30 & metadata$BMI > 25, "C",
                     ifelse(metadata$BMI > 30, "D", NA)))))

metadata$FEV1_group <- with(metadata,
                       ifelse(metadata$FEV1. > 0.80, "A",
                       ifelse(metadata$FEV1. > 0.50 & metadata$FEV1. <= 0.80, "B",
                       ifelse(metadata$FEV1. > 0.30 & metadata$FEV1. <= 0.50, "C",
                       ifelse(metadata$FEV1. <= 0.30, "D", NA)))))

metadata$hospital_group <- with(metadata,
                          ifelse(metadata$no_of_hospitalisations_for_resp_problems_exac_in_the_past_12_months > 0, "yes", "no"))

metadata$exacerbation_group <- with(metadata,
                               ifelse(metadata$No_of_exac_in_the_past_year == 0, "A",
                               ifelse(metadata$No_of_exac_in_the_past_year > 0 &
                                        metadata$No_of_exac_in_the_past_year <= 2, "B",
                               ifelse(metadata$No_of_exac_in_the_past_year >= 3, "C", NA))))


metadata$mrc_group <- with(metadata,
                      ifelse(metadata$MRC_dyspnoea_score < 4, "A",
                      ifelse(metadata$MRC_dyspnoea_score == 4, "B",
                      ifelse(metadata$MRC_dyspnoea_score == 5, "C", NA))))
#96xA vs 10xB

#metadata$MRC_dyspnoea_score <- as.factor(metadata$MRC_dyspnoea_score)



```

#statistical testing of clinical metadata
```{r}

#for this, delete longitudinal samples: Dundee007, Dundee034, Dundee036, Dundee039
metadata_single <- metadata[-c(7,34,36,39),]

#exclude Dundee048: only 173 Neisseria mucosa reads, although 14 mio sequencing reads (outlier in beta diversity analysis)
metadata_single <- metadata_single[-c(44),]

table(metadata_single$aetiology)
metadata_single$aetiology <- with(metadata_single,
                                  ifelse(aetiology == "ABPA", "ABPA", 
                                  ifelse(aetiology == "alpha-1 antitrypsin deficiency", 
                                         "alpha-1AntiTrypsinDeficiency",
                                  ifelse(aetiology == "Aspiration", "Aspiration",
                                  ifelse(aetiology == "asthma", "Asthma",
                                  ifelse(aetiology == "Asthma", "Asthma",
                                  ifelse(aetiology == "CFTRRD", "CFTR-RD",
                                  ifelse(aetiology == "copd", "COPD",
                                  ifelse(aetiology == "COPD", "COPD",
                                  ifelse(aetiology == "idiopathic", "idiopathic",
                                  ifelse(aetiology == "immunodeficiency", "immunodeficiency",
                                  ifelse(aetiology == "inflammatoryboweldisease",
                                         "InflammatoryBowelDisease",
                                  ifelse(aetiology == "NTM", "NTM",
                                  ifelse(aetiology == "PCD", "PCD",
                                  ifelse(aetiology == "postinfective", "post-infective",
                                  ifelse(aetiology == "Rheumatoid arthritis", "RheumatoidArthritis", 
                                         NA))))))))))))))))

table(metadata_single$BMI_group)
table(metadata_single$age_group)

##gender
kruskal.test(metadata_single$MRC_dyspnoea_score ~ metadata_single$sex)
#p=0.89
kruskal.test(metadata_single$mrc_group ~ metadata_single$sex)
#p=0.007

table(metadata_single$mrc_group)
#92x A - 9x B, B nur in Männern


kruskal.test(metadata_single$No_of_exac_in_the_past_year ~ metadata_single$sex)
#p=0.94
kruskal.test(metadata_single$exacerbation_group ~ metadata_single$sex)
#p=0.96


kruskal.test(metadata_single$no_of_hospitalisations_for_resp_problems_exac_in_the_past_12_months ~ metadata_single$sex)
#p=0.22
kruskal.test(metadata_single$hospital_group ~ metadata_single$sex)
#p=0.26


t.test(metadata_single$BMI ~ metadata_single$sex)
#p=0.07
kruskal.test(metadata_single$BMI_group ~ metadata_single$sex)
#p=0.07


t.test(metadata_single$age ~ metadata_single$sex)
#p=0.29
kruskal.test(metadata_single$age_group ~ metadata_single$sex)
#p=0.13



###commensals
metadata_single$group_commensal <- with(metadata_single,
                          ifelse(metadata_single$group == "commensals", "commensal",
                          ifelse(metadata_single$group == "PA", "other",
                          ifelse(metadata_single$group == "PA_SA", "other",
                          ifelse(metadata_single$group == "HI", "other",
                          ifelse(metadata_single$group == "HI_MC", "other",
                          ifelse(metadata_single$group == "MC", "other",
                          ifelse(metadata_single$group == "SA", "other",
                          ifelse(metadata_single$group == "Candida", "other",
                          ifelse(metadata_single$group == "Herpesvirus", "other",
                          ifelse(metadata_single$group == "SM", "other",
                          ifelse(metadata_single$group == "other", "other",
                          ifelse(metadata_single$group == "P_other", "other",
                          ifelse(metadata_single$group == "SL", "other", 
                          ifelse(metadata_single$group == "NM", "other", 
                          ifelse(metadata_single$group == "EF", "other", NA))))))))))))))))

table(metadata_single$group_commensal)


metadata_single %>%
  kruskal_test(group_commensal ~ BSI)
#p=0.022

metadata_single %>% 
  dunn_test(group_commensal ~ BSI, p.adjust.method = "BH")
 

aa <- subset(metadata_single, metadata_single$group_commensal == "commensal")
bb <- subset(metadata_single, metadata_single$group_commensal == "other")

count(aa, aa$bronchiectasis_severity_index)
count(bb, bb$bronchiectasis_severity_index)

###PA                       
metadata_single$group_PA <- with(metadata_single,
                          ifelse(metadata_single$group == "commensals", "other",
                          ifelse(metadata_single$group == "PA", "Pseudomonas",
                          ifelse(metadata_single$group == "PA_SA", "Pseudomonas",
                          ifelse(metadata_single$group == "HI", "other",
                          ifelse(metadata_single$group == "HI_MC", "other",
                          ifelse(metadata_single$group == "MC", "other",
                          ifelse(metadata_single$group == "SA", "other",
                          ifelse(metadata_single$group == "Candida", "other",
                          ifelse(metadata_single$group == "Herpesvirus", "other",
                          ifelse(metadata_single$group == "SM", "other",
                          ifelse(metadata_single$group == "other", "other",
                          ifelse(metadata_single$group == "P_other", "other",
                          ifelse(metadata_single$group == "SL", "other",
                          ifelse(metadata_single$group == "NM", "other", 
                          ifelse(metadata_single$group == "EF", "other", NA))))))))))))))))

table(metadata_single$group_PA)

#BSI
metadata_single %>%
  kruskal_test(group_PA ~ BSI)
#p=0.07

metadata_single %>% 
  dunn_test(group_PA ~ BSI, p.adjust.method = "BH")

aa <- subset(metadata_single, metadata_single$group_PA == "Pseudomonas")
bb <- subset(metadata_single, metadata_single$group_PA == "other")

count(aa, aa$BSI)
count(bb, bb$BSI)


#t test
t.test(metadata_single$FEV1. ~ metadata_single$group_PA)
#p=0.62

t.test(metadata_single$long_term_antibiotic ~ metadata_single$group_PA)
#p=0.048

t.test(metadata_single$Inhaled_steroid ~ metadata_single$group_PA)
#p=0.51


##HI
metadata_single$group_HI <- with(metadata_single,
                          ifelse(metadata_single$group == "commensals", "other",
                          ifelse(metadata_single$group == "PA", "other",
                          ifelse(metadata_single$group == "PA_SA", "other",
                          ifelse(metadata_single$group == "HI", "Haemophilus",
                          ifelse(metadata_single$group == "HI_MC", "Haemophilus",
                          ifelse(metadata_single$group == "MC", "other",
                          ifelse(metadata_single$group == "SA", "other",
                          ifelse(metadata_single$group == "Candida", "other",
                          ifelse(metadata_single$group == "Herpesvirus", "other",
                          ifelse(metadata_single$group == "SM", "other",
                          ifelse(metadata_single$group == "other", "other",
                          ifelse(metadata_single$group == "P_other", "other",
                          ifelse(metadata_single$group == "SL", "other",
                          ifelse(metadata_single$group == "NM", "other", 
                          ifelse(metadata_single$group == "EF", "other", NA))))))))))))))))

table(metadata_single$group_HI)

#BSI
metadata_single %>%
  kruskal_test(group_HI ~ BSI)
#p=0.57

metadata_single %>% 
  dunn_test(group_HI ~ BSI, p.adjust.method = "BH")


aa <- subset(metadata_single, metadata_single$group_HI == "Haemophilus")
bb <- subset(metadata_single, metadata_single$group_HI == "other")

count(aa, aa$BSI)
count(bb, bb$BSI)


#t test
t.test(metadata_single$FEV1. ~ metadata_single$group_HI)
#p=0.24

t.test(metadata_single$long_term_antibiotic ~ metadata_single$group_HI)
#p=0.31

t.test(metadata_single$Inhaled_steroid ~ metadata_single$group_HI)
#p=0.31



##remaining individual samples
metadata_individuals <- subset(metadata_single, group_commensal == "other" & group_HI == "other" &
                                 group_PA == "other")

metadata_single$groupShort <- with(metadata_single,
                            ifelse(group_PA == "Pseudomonas", "Pseudomonas",
                            ifelse(group_HI == "Haemophilus", "Haemophilus",
                            ifelse(group_commensal == "commensal", "Commensals", "individual"))))


BSI_distribution <- metadata_single[,c(1,6,42)]

ggplot(BSI_distribution, aes(x = BSI, fill = groupShort)) +
  geom_bar(stat = "count", position = position_dodge(0.93)) +
  scale_fill_manual(values = c("Commensals" = "lightskyblue",
                               "Haemophilus" = "gold",
                               "Pseudomonas" = "darkgreen",
                               "individual" = "grey")) +
  labs(x = "Bronchiectasis Severity Index", y = "Count") +
  coord_flip() +
  theme_pubr(border = TRUE)

ggplot(BSI_distribution, aes(x = groupShort, fill = BSI)) +
  geom_bar(stat = "count", position = position_dodge(0.93)) +
  scale_fill_manual(values = c("mild" = "#FED98E",
                               "moderate" = "#FE9929",
                               "severe" = "#CC4C02")) +
  labs(x = "Group", y = "Count") +
  coord_flip() +
  theme_pubr(border = TRUE)


BSI_distribution %>%
  kruskal_test(groupShort ~ BSI)

BSI_distribution %>% 
  dunn_test(groupShort ~ BSI, p.adjust.method = "BH")

BSI_distribution %>%
  kruskal_test(BSI ~ groupShort)

BSI_distribution %>%
  dunn_test(BSI ~ groupShort, p.adjust.method = "BH")

```


#load abundance table
```{r}
RPMM_BE <- read_delim("RPMM_wochenende.rep.us.raspir.merged_final.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

RPMM_BE <- data.frame(RPMM_BE)

rownames(RPMM_BE) <- RPMM_BE$organism
RPMM_BE$organism <- NULL

#exclude longitudinal samples Dundee007, Dundee034, Dundee036, Dundee039 & exclude Dundee048
RPMM_BE_single <- RPMM_BE[,-c(7,34,36,39,48)]

#####make relative abundance table
RPMM_BE_single_t <- data.frame(t(RPMM_BE_single))

BE_single_rel_t <- make_relative(as.matrix(RPMM_BE_single_t))
BE_single_rel_t <- data.frame(BE_single_rel_t)

BE_single_rel <- data.frame(t(BE_single_rel_t))

#sorting
BE_single_rel$rowsum <- rowSums(BE_single_rel)
sum <- sum(BE_single_rel$rowsum)
BE_single_rel$abundance <- (BE_single_rel$rowsum / sum) * 100
BE_single_rel <- BE_single_rel[order(BE_single_rel$abundance, decreasing = TRUE),]
BE_single_rel$cumsum <- cumsum(BE_single_rel$abundance)
#95% most abundant bis Bifidobacterium longum

BE_single_rel$cumsum <- NULL
BE_single_rel$abundance <- NULL
BE_single_rel$rowsum <- NULL

#abundance filtering: exclude every species where every column has a threshold below 1%
#abundance > 1% in at least one sample
top_abundant <- BE_single_rel[!apply(BE_single_rel<0.01, 1, all, na.rm=TRUE),]
#102 species

#prevalence filtering: species must appear in 10% of all samples
prev_filt_abund = 90

BE_prev <- BE_single_rel

n_prev = round(length(colnames(BE_prev)) * prev_filt_abund / 100)
BE_prev[BE_prev == 0] <- NA
BE_prev <- BE_prev[rowSums(is.na(BE_prev)) < n_prev, ]
BE_prev[is.na(BE_prev)] <- 0
#70 species

#combine abundance and prevalence filtered datasets
#110 species
BE_filt <- unique(rbind(BE_prev, top_abundant))

BE_t <- data.frame(t(BE_filt))

#make new relative abundance table to again obtain 100%
BE_t <- make_relative(as.matrix(BE_t))
BE_t <- data.frame(BE_t)

BE <- data.frame(t(BE_t))
BE <- BE[order(rowSums(BE), decreasing = TRUE),]

BE_t <- data.frame(t(BE))

```


#load healthy metadata and abundance table
```{r}

metadata_healthy <- read_delim("metadata_healthy.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)
metadata_healthy <- data.frame(metadata_healthy)

rownames(metadata_healthy) <- metadata_healthy$sample_name


RPMM_healthy <- read_delim("RPMM_healthy_wochenende.rep.us.raspir.merged.taxonomy_bearbeitet.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

RPMM_healthy <- data.frame(RPMM_healthy)

rownames(RPMM_healthy) <- RPMM_healthy$species
RPMM_healthy$species <- NULL


#test which samples are unique for RPMM and/or metadata
#test_healthy <- merge(RPMM_healthy_t, metadata_healthy, by = 0, 
#                      all = TRUE)
#rownames(test_healthy) <- test_healthy$Row.names
#test_healthy00 <- test_healthy[,c(1,161,162)]
##results: Sp_104&Sp_291 only in metadata, Sp_282 only as fastq

#metadata healthy: 157 samples, RPMM healthy
#exclude Sp_282 in metadata, exclude Sp_104 & Sp_291 in RPMM
metadata_healthy <- metadata_healthy[-c(2,36),]
RPMM_healthy$Sp_282 <- NULL

##add age group
metadata_healthy$ageGroup <- with(metadata_healthy,
                                  ifelse(metadata_healthy$age < 20, "A",
                                  ifelse(metadata_healthy$age >= 20 & age < 50, "B",
                                  ifelse(metadata_healthy$age >=50, "C", NA))))


metadata_healthy <- metadata_healthy[order(rownames(metadata_healthy)),]


#adapt filtering from bronchiectasis
#exclude people <20 years
#####make relative abundance table
RPMM_healthy_t <- data.frame(t(RPMM_healthy))
RPMM_healthy_t <- RPMM_healthy_t[order(rownames(RPMM_healthy_t)),]

RPMM_healthy_t$ageGroup <- metadata_healthy$ageGroup
RPMM_healthy_t <- subset(RPMM_healthy_t, ageGroup!="A")
RPMM_healthy_t$ageGroup <- NULL

metadata_healthy <- subset(metadata_healthy, ageGroup!="A")

#export new metadata
write.table(metadata_healthy,
            file = "metadata_healthy_filt20years.csv",
            col.names = TRUE, sep = ";")

#exclude smokers from healthy controls: Sp115, Sp155, Sp281, Sp320, Sp414, Sp416, Sp768, Sp941
smokersToRemove <- c("Sp_115", "Sp_155", "Sp_281", "Sp_320", "Sp_414", "Sp_416", 
                     "Sp_768", "Sp_941")

metadata_healthy <- metadata_healthy[!(row.names(metadata_healthy) %in% smokersToRemove),]
RPMM_healthy_t <- RPMM_healthy_t[!(row.names(RPMM_healthy_t) %in% smokersToRemove),]


healthy_rel_t <- make_relative(as.matrix(RPMM_healthy_t))
healthy_rel_t <- data.frame(healthy_rel_t)

healthy_rel <- data.frame(t(healthy_rel_t))

#sorting
#healthy_rel$rowsum <- rowSums(healthy_rel)
#sum <- sum(healthy_rel$rowsum)
#healthy_rel$abundance <- (healthy_rel$rowsum / sum) * 100
#healthy_rel <- healthy_rel[order(healthy_rel$abundance, decreasing = TRUE),]
#healthy_rel$cumsum <- cumsum(healthy_rel$abundance)
#95% most abundant bis Streptococcus anginosus

#healthy_rel$cumsum <- NULL
#healthy_rel$abundance <- NULL
#healthy_rel$rowsum <- NULL

#abundance filtering: exclude every species where every column has a threshold below 1%
#abundance > 1% in at least one sample
top_abundant_h <- healthy_rel[!apply(healthy_rel<0.01, 1, all, na.rm=TRUE),]
#70 species

#prevalence filtering: species must appear in 10% of all samples
prev_filt_abund = 90

healthy_prev <- healthy_rel

n_prev_h = round(length(colnames(healthy_prev)) * prev_filt_abund / 100)
healthy_prev[healthy_prev == 0] <- NA
healthy_prev <- healthy_prev[rowSums(is.na(healthy_prev)) < n_prev_h, ]
healthy_prev[is.na(healthy_prev)] <- 0
#87 species

#combine abundance and prevalence filtered datasets
#90 species
healthy_filt <- unique(rbind(healthy_prev, top_abundant_h))

healthy_t <- data.frame(t(healthy_filt))

#make new relative abundance table to again obtain 100%
healthy_t <- make_relative(as.matrix(healthy_t))
healthy_t <- data.frame(healthy_t)

healthy <- data.frame(t(healthy_t))
healthy <- healthy[order(rowSums(healthy), decreasing = TRUE),]

healthy_t <- data.frame(t(healthy))

##calculate healthy cumulative dataset
healthyCumulative <- data.frame(rowMeans(healthy))
colnames(healthyCumulative) <- c("healthy")

```


#make relative abundance table with healthy included
```{r}
#combine BE and cumulative healthy

BE_healthyCumulative <- merge(healthyCumulative, BE, all = TRUE, by = 0)
rownames(BE_healthyCumulative) <- BE_healthyCumulative$Row.names
BE_healthyCumulative$Row.names <- NULL
BE_healthyCumulative[is.na(BE_healthyCumulative)] <- 0

BE_healthyCumulative <- BE_healthyCumulative[order(rowSums(BE_healthyCumulative), 
                                                   decreasing = TRUE),]
BE_healthyCumulative_t <- data.frame(t(BE_healthyCumulative))

metadata_BE_healthyCumulative <- metadata_single
metadata_BE_healthyCumulative <- metadata_BE_healthyCumulative[,c(1,2,3,6,8,9)]

#add healthy (mean age 35 years)
metadata_BE_healthyCumulative[nrow(metadata_BE_healthyCumulative) + 1,] <- 
  c("healthy", "healthy", "healthy", "healthy", "healthy", 35)

rownames(metadata_BE_healthyCumulative) <- metadata_BE_healthyCumulative$Lab_ID
metadata_BE_healthyCumulative$state <- c(rep("Bronchiectasis", 101),
                                         "Healthy")

#combine BE and all healthy samples
BE_healthy <- merge(healthy, BE, all = TRUE, by = 0)
rownames(BE_healthy) <- BE_healthy$Row.names
BE_healthy$Row.names <- NULL
BE_healthy[is.na(BE_healthy)] <- 0

BE_healthy <- BE_healthy[order(rowSums(BE_healthy), decreasing = TRUE),]
BE_healthy_t <- data.frame(t(BE_healthy))

#make small metadata tables
mBE <- metadata_single
mBE00 <- mBE[,c(1,8,9,18)]  
colnames(mBE00) <- c("sample", "gender", "age", "BMI")  

mHealthy <- metadata_healthy
mHealthy00 <- mHealthy[,c(1,3,4,7)]
colnames(mHealthy00) <- c("sample", "gender", "age", "BMI")
mHealthy00$gender <- with(mHealthy00,
                          ifelse(gender == "m", "male",
                          ifelse(gender == "f", "female", NA)))

metadata_BE_healthy <- rbind(mBE00, mHealthy00)
metadata_BE_healthy$state <- c(rep("Bronchiectasis", 101),
                               rep("Healthy", 88))

#export tables for MaAsLin2 analysis
write.csv(BE_healthy, file = "BE_healthy_R.csv", row.names = TRUE)
write.csv(metadata_BE_healthy, file = "metadata_BE_healthy_R.csv", row.names = TRUE,
            col.names = TRUE)

```

#include healthy smokers
```{r}
RPMM_healthySmokers <- read_delim("RPMM_healthysmokers.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

RPMM_healthySmokers <- data.frame(RPMM_healthySmokers)
rownames(RPMM_healthySmokers) <- RPMM_healthySmokers$species
RPMM_healthySmokers$species <- NULL

#filter dataset like it BE and healthy datasets
#####make relative abundance table
RPMM_healthySmokers_t <- data.frame(t(RPMM_healthySmokers))

smokers_rel_t <- make_relative(as.matrix(RPMM_healthySmokers_t))
smokers_rel_t <- data.frame(smokers_rel_t)

smokers_rel <- data.frame(t(smokers_rel_t))

#abundance filtering: exclude every species where every column has a threshold below 1%
#abundance > 1% in at least one sample
top_abundant_s <- smokers_rel[!apply(smokers_rel<0.01, 1, all, na.rm=TRUE),]
#33 species

#prevalence filtering: species must appear in 10% of all samples
prev_filt_abund = 90

smokers_prev <- smokers_rel

n_prev = round(length(colnames(smokers_prev)) * prev_filt_abund / 100)
smokers_prev[smokers_prev == 0] <- NA
smokers_prev <- smokers_prev[rowSums(is.na(smokers_prev)) < n_prev, ]
smokers_prev[is.na(smokers_prev)] <- 0
#87 species

#combine abundance and prevalence filtered datasets
#88 species
smokers_filt <- unique(rbind(top_abundant_s, smokers_prev))

smokers_t <- data.frame(t(smokers_filt))

#make new relative abundance table to again obtain 100%
smokers_t <- make_relative(as.matrix(smokers_t))
smokers_t <- data.frame(smokers_t)

smokers <- data.frame(t(smokers_t))
smokers <- smokers[order(rowSums(smokers), decreasing = TRUE),]

smokers_t <- data.frame(t(smokers))

```

#combine BE, healthy and smoker filtered datasets
```{r}

#merge data sets
BE_healthy_hSmokers <- merge(BE_healthy, smokers, all = TRUE, by = 0)
rownames(BE_healthy_hSmokers) <- BE_healthy_hSmokers$Row.names
BE_healthy_hSmokers$Row.names <- NULL
BE_healthy_hSmokers[is.na(BE_healthy_hSmokers)] <- 0

BE_healthy_hSmokers <- BE_healthy_hSmokers[order(rowSums(BE_healthy_hSmokers), decreasing = TRUE),]
BE_healthy_hSmokers_t <- data.frame(t(BE_healthy_hSmokers))

#make BE, healthy cumulative and all smoker dataset
BE_healthyCumulative_smokers <- merge(BE_healthyCumulative, smokers, all = TRUE, by = 0)
rownames(BE_healthyCumulative_smokers) <- BE_healthyCumulative_smokers$Row.names
BE_healthyCumulative_smokers$Row.names <- NULL
BE_healthyCumulative_smokers[is.na(BE_healthyCumulative_smokers)] <- 0

BE_healthyCumulative_smokers <- BE_healthyCumulative_smokers[order(rowSums(
  BE_healthyCumulative_smokers), decreasing = TRUE),]

BE_healthyCumulative_smokers_t <- data.frame(t(BE_healthyCumulative_smokers))

#basic metadata BE, healthy all and healthy smoker
mSmo <- data.frame(sample = c("EGLA00101", "EGLA017", "EGLA018", "EGLA022", "EGLA048", 
                                    "EGLA052", "EGLA055", "EGLA058"),
                   state = c(rep("Healthy smoker", 8)))
rownames(mSmo) <- mSmo$sample

metadata_BE_healthyall_hSmokers <- rbind(metadata_BE_healthy[,c(1,5)], mSmo)

#basic metadata BE, healthy cumulative and healthy smoker
metadata_BE_healthyCum_hSmokers <- metadata_BE_healthyCumulative
metadata_BE_healthyCum_hSmokers <- metadata_BE_healthyCum_hSmokers[, -c(5,6)]

metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA00101", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA017", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA018", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA022", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA048", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA052", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA055", "smoker", "smoker", "smoker", "Healthy smoker")
metadata_BE_healthyCum_hSmokers[nrow(metadata_BE_healthyCum_hSmokers) + 1, ] <-
  c("EGLA058", "smoker", "smoker", "smoker", "Healthy smoker")

rownames(metadata_BE_healthyCum_hSmokers) <- metadata_BE_healthyCum_hSmokers$Lab_ID

```


#alpha diversity
```{r}

metadata_single$shannon <- diversity(BE_t, index = "shannon")
metadata_single$simpson <- diversity(BE_t, index = "simpson")
metadata_single$specNumber <- specnumber(BE_t)
metadata_single$richness <- metadata_single$specNumber / sqrt(length(metadata_single$specNumber))
metadata_single$evenness <- metadata_single$shannon / log(metadata_single$specNumber)


metadata_single %>% 
  kruskal_test(shannon ~ BSI)
#p=0.002
metadata_single %>%
  kruskal_effsize(shannon ~ BSI, ci = TRUE)

metadata_single %>% 
  dunn_test(shannon ~ BSI, p.adjust.method = "BH")



metadata_single %>% 
  kruskal_test(simpson ~ BSI)
#p=0.008
metadata_single %>%
  kruskal_effsize(simpson ~ BSI, ci = TRUE)

metadata_single %>% 
  kruskal_test(specNumber ~ BSI)
#p=0.005
metadata_single %>%
  kruskal_effsize(specNumber ~ BSI, ci = TRUE)

metadata_single %>% 
  kruskal_test(evenness ~ BSI)
#p=0.02
metadata_single %>%
  kruskal_effsize(evenness ~ BSI, ci = TRUE)

#stat.test1 <- compare_means(data = metadata_single, shannon ~ BSI,
                            #method = "wilcox.test")
stat.test1 <- metadata_single %>%
  dunn_test(shannon ~ BSI, p.adjust.method = "BH") %>%
  add_xy_position(x = "BSI")

stat.test1

ggplot(metadata_single, aes(x = BSI, y = shannon)) +
  geom_violin(width = 0.5) +
  geom_point() +
  theme_pubr() +
  xlab("Bronchiectasis Severity Index") + ylab("Shannon diversity") +
  #stat_compare_means(method = "kruskal.test", label.x.npc = "left") +
  stat_pvalue_manual(stat.test1, label = "p.adj.signif",
                     tip.length = 0.01) +
  geom_pointrange(mapping = aes(x = BSI, y = shannon),
                  stat = "summary",
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median,
                  colour = "red", size = 0.3)

ggplot(metadata_single, aes(x = BSI, y = shannon)) +
  geom_violin(aes(fill = BSI, fill = after_scale(colorspace::lighten(fill, 0.5)))) +
  geom_boxplot(fill = "white", width = 0.2, outlier.shape = NA, coef = 0) +
  geom_point(aes(color = BSI), position = position_jitter(width = 0.03, seed = 0),
             alpha = 0.5) +
  geom_point(color = "black", position = position_jitter(width = 0.03, seed = 0),
             stroke = 0.7, shape = 1) +
  scale_fill_manual(values = c("mild" = "#FED98E",
                               "moderate" = "#FE9929",
                               "severe" = "#CC4C02")) +
  scale_color_manual(values = c("mild" = "#FED98E",
                                "moderate" = "#FE9929", 
                                "severe" = "#CC4C02")) +
  stat_pvalue_manual(stat.test1, label = "p.adj.signif", tip.length = 0.01, hide.ns = TRUE) +
  theme_pubr(border = TRUE, legend = "none") +
  labs(x = "Bronchiectasis Severity Index", y = "Shannon diversity")


stat.test2 <- metadata_single %>%
  dunn_test(simpson ~ BSI, p.adjust.method = "BH") %>%
  add_xy_position(x = "BSI")

stat.test2

ggplot(metadata_single, aes(x = BSI, y = simpson)) +
  geom_violin(aes(fill = BSI, fill = after_scale(colorspace::lighten(fill, 0.5)))) +
  geom_boxplot(fill = "white", width = 0.12, outlier.shape = NA, coef = 0) +
  geom_point(aes(color = BSI), position = position_jitter(width = 0.03, seed = 0),
             alpha = 0.5) +
  geom_point(color = "black", position = position_jitter(width = 0.03, seed = 0),
             stroke = 0.7, shape = 1) +
  scale_fill_manual(values = c("mild" = "#FED98E",
                               "moderate" = "#FE9929",
                               "severe" = "#CC4C02")) +
  scale_color_manual(values = c("mild" = "#FED98E",
                                "moderate" = "#FE9929", 
                                "severe" = "#CC4C02")) +
  stat_pvalue_manual(stat.test2, label = "p.adj.signif", tip.length = 0.01, hide.ns = TRUE,
                     y.position = c(1, 1.05)) +
  theme_pubr(border = TRUE, legend = "none") +
  labs(x = "Bronchiectasis Severity Index", y = "Simpson diversity index")


stat.test3 <- metadata_single %>%
  dunn_test(specNumber ~ BSI, p.adjust.method = "BH") %>%
  add_xy_position(x = "BSI")

stat.test3

ggplot(metadata_single, aes(x = BSI, y = specNumber)) +
  geom_violin(aes(fill = BSI, fill = after_scale(colorspace::lighten(fill, 0.5)))) +
  geom_boxplot(fill = "white", width = 0.3, outlier.shape = NA, coef = 0) +
  geom_point(aes(color = BSI), position = position_jitter(width = 0.03, seed = 0),
             alpha = 0.5) +
  geom_point(color = "black", position = position_jitter(width = 0.03, seed = 0),
             stroke = 0.7, shape = 1) +
  scale_fill_manual(values = c("mild" = "#FED98E",
                               "moderate" = "#FE9929",
                               "severe" = "#CC4C02")) +
  scale_color_manual(values = c("mild" = "#FED98E",
                                "moderate" = "#FE9929", 
                                "severe" = "#CC4C02")) +
  stat_pvalue_manual(stat.test3, label = "p.adj.signif", tip.length = 0.01, hide.ns = TRUE,
                     y.position = c(61,64)) +
  theme_pubr(border = TRUE, legend = "none") +
  labs(x = "Bronchiectasis Severity Index", y = "Species number")


stat.test4 <- metadata_single %>%
  dunn_test(evenness ~ BSI, p.adjust.method = "BH") %>%
  add_xy_position(x = "BSI")

stat.test4

ggplot(metadata_single, aes(x = BSI, y = evenness)) +
  geom_violin(aes(fill = BSI, fill = after_scale(colorspace::lighten(fill, 0.5)))) +
  geom_boxplot(fill = "white", width = 0.2, outlier.shape = NA, coef = 0) +
  geom_point(aes(color = BSI), position = position_jitter(width = 0.03, seed = 0),
             alpha = 0.5) +
  geom_point(color = "black", position = position_jitter(width = 0.03, seed = 0),
             stroke = 0.7, shape = 1) +
  scale_fill_manual(values = c("mild" = "#FED98E",
                               "moderate" = "#FE9929",
                               "severe" = "#CC4C02")) +
  scale_color_manual(values = c("mild" = "#FED98E",
                                "moderate" = "#FE9929", 
                                "severe" = "#CC4C02")) +
  scale_y_continuous(limits = c(0,1)) +
  stat_pvalue_manual(stat.test4, label = "p.adj.signif", tip.length = 0.01, hide.ns = TRUE,
                     y.position = c(0.93, 0.98)) +
  theme_pubr(border = TRUE, legend = "none") +
  labs(x = "Bronchiectasis Severity Index", y = "Pielou's evenness index")

#test metadata in commensal/Haemophilus/Pseudomonas groups

metadata_HI <- subset(metadata_single, metadata_single$group00 == "Haemophilus")

metadata_HI %>% 
  kruskal_test(shannon ~ BSI)
#p=0.01
metadata_HI %>%
  dunn_test(shannon ~ BSI, p.adjust.method = "BH")

metadata_HI %>%
  kruskal_effsize(shannon ~ BSI, ci = TRUE)

#make figure
metadata_single$groupAlpha <- with(metadata_single,
                                    ifelse(group00 == "Pseudomonas", "Pseudomonas",
                                    ifelse(group00 == "Haemophilus", "Haemophilus",
                                    ifelse(group00 == "commensals", "commensals", 
                                           "rare pathogens"))))

subset(metadata_single, groupAlpha == "Haemophilus") %>% 
  kruskal_test(shannon ~ BSI)
subset(metadata_single, groupAlpha == "Haemophilus") %>% 
  kruskal_effsize(shannon ~ BSI, ci = TRUE)


subset(metadata_single, groupAlpha == "Pseudomonas") %>% 
  kruskal_test(shannon ~ BSI)

subset(metadata_single, groupAlpha == "commensals") %>% 
  kruskal_test(shannon ~ BSI)
subset(metadata_single, groupAlpha == "commensals") %>% 
  kruskal_effsize(shannon ~ BSI, ci = TRUE)

subset(metadata_single, groupAlpha == "rare pathogens") %>% 
  kruskal_test(shannon ~ BSI)
subset(metadata_single, groupAlpha == "rare pathogens") %>% 
  kruskal_effsize(shannon ~ BSI, ci = TRUE)

ggplot(metadata_single, aes(x = BSI, y = shannon)) +
  geom_boxplot(width = 0.5) +
  geom_point() +
  facet_grid( ~ groupAlpha) +
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, hide.ns = TRUE) +
  ylab("Shannon diversity") + xlab("Bronchiectasis Severity Index") +
  theme_pubr(border = TRUE)



```

#pie chart with pathogen distribution
```{r}
metadata_single$pieChart <- with(metadata_single,
                          ifelse(group == "commensals", "Commensals",
                          ifelse(group == "PA", "Pseudomonas",
                          ifelse(group == "HI", "Haemophilus",
                          ifelse(group == "Candida", "Candida",
                          ifelse(group == "HI_MC", "Haemophilus",
                          ifelse(group == "PA_SA", "Pseudomonas",
                          ifelse(group == "SA", "Staphylococcus",
                          ifelse(group == "P_other", "Pseudomonas",
                          ifelse(group == "MC", "Moraxella",
                          ifelse(group == "Herpesvirus", "Herpesvirus",
                          ifelse(group == "SM", "Stenotrophomonas",
                          ifelse(group == "NM", "Neisseria",
                          ifelse(group == "SL", "Serratia",
                          ifelse(group == "other", "Enterococcus", 
                          ifelse(group == "EF", "Enterococcus", NA))))))))))))))))

table(metadata_single$group)

ggplot(metadata_single, aes(x = "", y = pieChart, fill = pieChart)) +
  geom_bar(stat = "identity") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = c("Commensals" = "lightskyblue",
                               "Candida" = "firebrick2",
                               "Herpesvirus" = "orange",
                               "Haemophilus" = "gold",
                               "Moraxella" = "goldenrod4",
                               "Neisseria" = "palevioletred",
                               "Pseudomonas" = "darkgreen",
                               "Serratia" = "yellowgreen",
                               "Staphylococcus" = "mediumblue",
                               "Stenotrophomonas" = "seagreen")) +
  theme_void()

#get frequencies
major_pathogens <- as.data.frame(table(metadata_single$pieChart))
sum_freq <- sum(major_pathogens$Freq)
major_pathogens$frequency <- round((major_pathogens$Freq / sum_freq) * 100)

df2 <- major_pathogens %>%
  mutate(csum = rev(cumsum(rev(Freq))),
         pos = Freq/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Freq/2, pos))

ggplot(df2, aes(x = "", y = Freq, fill = Var1)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Freq, "%")),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  scale_fill_manual(values = c("Commensals" = "lightskyblue",
                               "Candida" = "firebrick2",
                               "Herpesvirus" = "orange",
                               "Haemophilus" = "gold",
                               "Moraxella" = "goldenrod4",
                               "Neisseria" = "palevioletred",
                               "Pseudomonas" = "darkgreen",
                               "Serratia" = "yellowgreen",
                               "Staphylococcus" = "royalblue4",
                               "Stenotrophomonas" = "seagreen")) +
  guides(fill = guide_legend(title = "Group")) +
  theme_void()

#treemap
mycolors = c("#EE2C2C", "#87CEFA", "#CD661D", "#FFD700", "#FFA500",  "#8B6914", "#DB7093", "#006400",
             "#9ACD32", "#27408B", "#2E8B57")

col2hex("lightskyblue")
col2hex("firebrick2")
col2hex("orange")
col2hex("gold")
col2hex("goldenrod4")
col2hex("palevioletred")
col2hex("darkgreen")
col2hex("yellowgreen")
#col2hex("mediumblue")
col2hex("royalblue4")
col2hex("seagreen")
col2hex("chocolate3")

df2$mycolors <- mycolors
df2$percent <- c("7%", "38%", "2%", "27%", "1%", "1%", "1%", "15%", "1%", "5%", "3%")

treemap(df2,
        index = c("Var1", "percent"),
        vSize = "frequency",
        type = "color",
        vColor = "mycolors",
        fontsize.labels = c(11,10),
        fontcolor.labels = c("black", "black"),
        align.labels = list(
          c("center", "center"),
          c("right", "bottom")
        ),
        border.lwds = c(1, 1),
        inflate.labels = FALSE,
        title = "")

```


#beta diversity
```{r}

#NMDS

set.seed(2)

mds <- metaMDS(BE_t, distance = "bray", k = 3, autotransform = TRUE, trymax = 100 )
#stress=0.15

stressplot(mds)

data.scores <- as.data.frame(scores(mds)$sites)

metadata_single <- metadata_single[order(rownames(metadata_single)),]

data.scores$gender <- metadata_single$sex
data.scores$group <- metadata_single$group
data.scores$group00 <- metadata_single$group00
data.scores$BSI <- metadata_single$BSI
data.scores$FEV1_group <- metadata_single$FEV1_group
data.scores$MRC_group <- metadata_single$mrc_group
data.scores$exacerbation_group <- metadata_single$exacerbation_group
data.scores$hospital_group <- metadata_single$hospital_group
data.scores$age_group <- metadata_single$age_group
data.scores$pieChart <- metadata_single$pieChart
data.scores$shannon <- diversity(BE_t, index = "shannon")
data.scores$aetiology <- metadata_single$aetiology

envit_metadata <- data.scores
envit_metadata$NMDS1 <- NULL
envit_metadata$NMDS2 <- NULL
envit_metadata$NMDS3 <- NULL

envit_metadata <- data.frame(cbind(envit_metadata, BE_t))

data.envit <- envfit(mds, envit_metadata, permutation = 999, 
                             na.rm = TRUE)
data.envit


p1 <- ggplot(data.scores, aes(x = NMDS1, y = NMDS2, color = group00)) +
  geom_point(size = 2) +
  #stat_ellipse(geom = "polygon", alpha = 0.1) +
  scale_color_manual(values = c("commensals" = "lightskyblue",
                               "Candida" = "firebrick2",
                               #"Herpesvirus" = "orange",
                               "Herpesvirus" = "plum",
                               #"Haemophilus" = "gold",
                               "Haemophilus" = "gold3",
                               "Moraxella" = "goldenrod4",
                               #"Neisseria" = "palevioletred",
                               "Neisseria" = "maroon4",
                               "Pseudomonas" = "darkgreen",
                               "Serratia" = "yellowgreen",
                               "Staphylococcus" = "royalblue4",
                               #"Stenotrophomonas" = "seagreen",
                               "Stenotrophomonas" = "seagreen3",
                               "Enterococcus" = "chocolate3"),
                     name = "Group") +
  theme_pubr(border = TRUE, legend = "right") 

p1

#######
##add density plot

xdens <- 
  axis_canvas(p1, axis = "x") +
  geom_density(data = subset(data.scores, group00 == c("commensals", "Haemophilus", "Pseudomonas")), 
               aes(x = NMDS1, fill = group00), alpha = 0.6) +
  scale_fill_manual(values = c("commensals" = "lightskyblue",
                               "Haemophilus" = "gold",
                               "Pseudomonas" = "darkgreen"))

ydens <- 
  axis_canvas(p1, axis = "y", coord_flip = TRUE) +
  geom_density(data = subset(data.scores, group00 == c("commensals", "Haemophilus", "Pseudomonas")), 
               aes(x = NMDS2, fill = group00), alpha = 0.6) +
  scale_fill_manual(values = c("commensals" = "lightskyblue",
                               "Haemophilus" = "gold",
                               "Pseudomonas" = "darkgreen")) +
  coord_flip()

p1 %>%
  insert_xaxis_grob(xdens, grid::unit(1, "in"), position = "top") %>%
  insert_yaxis_grob(ydens, grid::unit(1, "in"), position = "right") %>%
  ggdraw()


###
#colour points based on various parameters
ggplot(data.scores, aes(x = NMDS1, y = NMDS2, color = BSI)) +
  geom_point(size = 2) +
  #stat_ellipse(geom = "polygon", alpha = 0.1) +
  scale_color_manual(values = c("#FED98E", "#FE9929", "#CC4C02"),
                     name = "Bronchiectasis\nSeverity Index") +
  theme_pubr(border = TRUE, legend = "right") 


ggplot(data.scores, aes(x = NMDS1, y = NMDS2, color = shannon)) +
  geom_point(size = 2) +
  scale_color_gradientn(colors = c("#BDC9E1", "#74A9CF", "#0570B0"),
                        name = "Shannon diversity") +
  theme_pubr(border = TRUE, legend = "right")

#calculate distances from each point to group centroid
vegdist_object <- vegdist(BE_t, method="bray", binary=FALSE)

group_vector <- factor(metadata_single$BSI)

betadisper_result <-
  betadisper(vegdist_object, group_vector, 
             type = c("centroid"), 
             bias.adjust = FALSE, sqrt.dist = FALSE, add = FALSE)

betadisper_df_distances <- 
  data.frame(value=betadisper_result$distances,
             group=group_vector)

stat.test <- betadisper_df_distances %>%
  wilcox_test(value ~ group) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p")
stat.test <- stat.test %>%
  add_xy_position(x = "group", fun = "max")
stat.test

betadisper_df_distances %>%
  kruskal_test(value ~ group)

betadisper_df_distances %>%
  kruskal_effsize(value ~ group, ci = TRUE)

ggplot(betadisper_df_distances, aes(x=group, y=value)) +
  geom_boxplot(width = 0.5, aes(fill = group)) +
  geom_point() +
  theme_pubr(border = TRUE) +
  scale_fill_manual(values = c("#FED98E", "#FE9929", "#CC4C02"),
                     name = "Bronchiectasis Severity Index") +
  stat_pvalue_manual(stat.test, label = "p.adj.signif", hide.ns = TRUE,
                     step.increase = 0.05, tip.length = 0.01) +
  ylab("Distance to centroid") +
  xlab("Bronchiectasis Severity Index") +
  theme(legend.position = "none")


#repeat clustering with cumulative healthy included

#NMDS

set.seed(2)

metadata_BE_healthyCumulative <- metadata_BE_healthyCumulative[order(rownames(
  metadata_BE_healthyCumulative)),]
BE_healthyCumulative_t <- BE_healthyCumulative_t[order(rownames(BE_healthyCumulative_t)),]


mds_healthyCumulative <- metaMDS(BE_healthyCumulative_t, distance = "bray", k = 3, 
                                 autotransform = TRUE, trymax = 100 )
#stress=0.15

stressplot(mds_healthyCumulative)

data.scores_healthyCumulative <- as.data.frame(scores(mds_healthyCumulative)$sites)



data.scores_healthyCumulative$age <- metadata_BE_healthyCumulative$age
data.scores_healthyCumulative$group <- metadata_BE_healthyCumulative$group00
data.scores_healthyCumulative$shannon <- diversity(BE_healthyCumulative_t, index = "shannon")

envit_metadata_healthyCumulative <- data.scores_healthyCumulative
envit_metadata_healthyCumulative$NMDS1 <- NULL
envit_metadata_healthyCumulative$NMDS2 <- NULL
envit_metadata_healthyCumulative$NMDS3 <- NULL

envit_metadata_healthyCumulative <- data.frame(cbind(envit_metadata_healthyCumulative, 
                                                     BE_healthyCumulative_t))

data.envit_healthyCumulative <- envfit(mds_healthyCumulative, envit_metadata_healthyCumulative,
                                       permutation = 999, na.rm = TRUE)
data.envit_healthyCumulative

data.scores_healthyCumulative$overall <- c(rep("BE", 101), "healthy")

data.scores_healthyCumulative$NMDS2 <- data.scores_healthyCumulative$NMDS2 * -1

p2 <- ggplot(data.scores_healthyCumulative, aes(x = NMDS1, y = NMDS2, color = group, 
                                          shape = overall)) +
  geom_point(aes(size = overall)) +
  #stat_ellipse(geom = "polygon", alpha = 0.1) +
  scale_color_manual(values = c("commensals" = "lightskyblue",
                               "Candida" = "firebrick2",
                               #"Herpesvirus" = "orange",
                               "Herpesvirus" = "plum",
                               #"Haemophilus" = "gold",
                               "Haemophilus" = "gold3",
                               "Moraxella" = "goldenrod4",
                               #"Neisseria" = "palevioletred",
                               "Neisseria" = "maroon4",
                               "Pseudomonas" = "darkgreen",
                               "Serratia" = "yellowgreen",
                               "Staphylococcus" = "royalblue4",
                               #"Stenotrophomonas" = "seagreen",
                               "Stenotrophomonas" = "seagreen3",
                               "Enterococcus" = "chocolate3",
                               "healthy" = "turquoise"),
                     breaks = c("healthy", "Candida", "Herpesvirus", "commensals",
                                "Enterococcus", 
                                "Haemophilus", "Moraxella", "Neisseria", "Pseudomonas",
                                "Serratia",
                                "Staphylococcus", "Stenotrophomonas"),
                     labels = c("Lung health", "Candida", "Herpesvirus", "Commensals", 
                                "Enterococcus", "Haemophilus", "Moraxella", "Neisseria",
                                "Pseudomonas", "Serratia", "Staphylococcus",
                                "Stenotrophomonas"),
                     name = "Group") +
  scale_size_manual(values = c(2, 3)) +
  theme_pubr(border = TRUE, legend = "right") +
  guides(shape = "none", size = "none")

p2

xdens_p2 <- 
  axis_canvas(p2, axis = "x") +
  geom_density(data = subset(data.scores_healthyCumulative, 
                             group == c("commensals", "Haemophilus", "Pseudomonas")), 
               aes(x = NMDS1, fill = group), alpha = 0.6) +
  scale_fill_manual(values = c("commensals" = "lightskyblue",
                               "Haemophilus" = "gold3",
                               "Pseudomonas" = "darkgreen"))

ydens_p2 <- 
  axis_canvas(p2, axis = "y", coord_flip = TRUE) +
  geom_density(data = subset(data.scores_healthyCumulative, 
                             group == c("commensals", "Haemophilus", "Pseudomonas")), 
               aes(x = NMDS2, fill = group), alpha = 0.6) +
  scale_fill_manual(values = c("commensals" = "lightskyblue",
                               "Haemophilus" = "gold3",
                               "Pseudomonas" = "darkgreen")) +
  coord_flip()

p2 %>%
  insert_xaxis_grob(xdens_p2, grid::unit(1, "in"), position = "top") %>%
  insert_yaxis_grob(ydens_p2, grid::unit(1, "in"), position = "right") %>%
  ggdraw()

#repeat clustering with all healthy and all healthy smokers included

#NMDS

set.seed(2)

metadata_BE_healthyall_hSmokers <- metadata_BE_healthyall_hSmokers[order(rownames(
  metadata_BE_healthyall_hSmokers)),]
BE_healthy_hSmokers_t <- BE_healthy_hSmokers_t[order(rownames(BE_healthy_hSmokers_t)),]


mds_healthyall_smoker <- metaMDS(BE_healthy_hSmokers_t, distance = "bray", k = 3, 
                                 autotransform = TRUE, trymax = 100 )
#stress=0.14

stressplot(mds_healthyall_smoker)

data.scores_healthyall_smoker <- as.data.frame(scores(mds_healthyall_smoker)$sites)



data.scores_healthyall_smoker$state <- metadata_BE_healthyall_hSmokers$state
data.scores_healthyall_smoker$sample <- rownames(data.scores_healthyall_smoker)

envit_metadata_healthyall_smoker <- data.scores_healthyall_smoker
envit_metadata_healthyall_smoker$NMDS1 <- NULL
envit_metadata_healthyall_smoker$NMDS2 <- NULL
envit_metadata_healthyall_smoker$NMDS3 <- NULL

envit_metadata_healthyall_smoker <- data.frame(cbind(envit_metadata_healthyall_smoker, 
                                                     BE_healthy_hSmokers_t))

data.envit_healthyall_smoker <- envfit(mds_healthyall_smoker, envit_metadata_healthyall_smoker,
                                       permutation = 999, na.rm = TRUE)
data.envit_healthyall_smoker


ggplot(data.scores_healthyall_smoker, aes(x = NMDS1, y = NMDS2, color = state, 
                                          shape = state)) +
  scale_x_continuous(limits = c(-2.4, 1.8)) +
  scale_color_discrete(labels = c("Bronchiectasis", "Healthy non-smoker", "Healthy smoker")) +
  scale_shape_discrete(labels = c("Bronchiectasis", "Healthy non-smoker", "Healthy smoker")) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = sample), size = 2) +
  stat_ellipse(geom = "polygon", alpha = 0.1, aes(fill = state)) +
  theme_pubr(border = TRUE, legend = "none") 

#draw vectors for most significant species
scores_env <- as.data.frame(scores(data.envit_healthyall_smoker, display = "vectors"))
scores_env <- cbind(scores_env, factor = rownames(scores_env))
scores_env$r <- data.envit_healthyall_smoker[["vectors"]][["r"]]
scores_env$p <- data.envit_healthyall_smoker[["vectors"]][["pvals"]]

scores_signif <- subset(scores_env, p == 0.001)

#scores_signif <- scores_signif[]

#shorten species name
scores_signif$species <- c("S. parasanguinis", "R. mucilaginosa", "H. influenzae", 
                           "P. melaninogenica", "S. salivarius", "V. atypica", "P. aeruginosa",
                           "N. subflava", "G. sanguinis", "V. dispar",
                           "P. jejuni", "S. odontolytica", "H. parainfluenzae", "E. sulci",
                           "F. periodonticum", "S. thermophilus",
                           "A. parvulum", "N. mucosa",
                           "L. mirabilis", "L. buccalis", "A. meyeri", "C. concisus", 
                           "L. oris")

scores_signif00 <- subset(scores_signif, r > 0.1)
scores_signif01 <- subset(scores_signif, r > 0.15)

ggplot(data.scores_healthyall_smoker) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, color = state),
             size = 3, alpha = 0.5) +
  scale_x_continuous(limits = c(-2.4, 1.8)) +
  coord_fixed() +
  geom_segment(data = scores_signif01,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), color = "grey") +
  geom_text_repel(data = scores_signif01, aes(x = NMDS1, y = NMDS2, label = species),
            size = 2.5, fontface = "italic") + 
  theme_pubr(border = TRUE, legend = "none")

options(ggrepel.max.overlaps = Inf)

```



#DMM clustering
```{r}

library(DirichletMultinomial)
library(reshape2)

#https://microbiome.github.io/tutorials/DMM.html

#RPMM_BE_single_t
kept_taxa <- names(BE_t)

RPMM_BE_single_t_filt <- RPMM_BE_single_t[,kept_taxa]

set.seed(2)

dmn(as.matrix(RPMM_BE_single_t_filt), k = 6, verbose = FALSE, 
    seed = runif(1, 0, .Machine$integer.max))

all_dmns <- 6
dmn_list <- numeric(all_dmns)


for(i in 1:all_dmns){
  print(i)
  assign(paste0("dmn_", i), dmn(as.matrix(RPMM_BE_single_t_filt), i, verbose = FALSE))
}

dmn_list <- list(dmn_1, dmn_2, dmn_3, dmn_4, dmn_5, dmn_6)

#save clusters / metacommunities
Dirichlet_multinomial_1 <- mixture(dmn_list[[1]], assign = TRUE)
Dirichlet_multinomial_2 <- mixture(dmn_list[[2]], assign = TRUE)
Dirichlet_multinomial_3 <- mixture(dmn_list[[3]], assign = TRUE)
Dirichlet_multinomial_4 <- mixture(dmn_list[[4]], assign = TRUE)
Dirichlet_multinomial_5 <- mixture(dmn_list[[5]], assign = TRUE)
Dirichlet_multinomial_6 <- mixture(dmn_list[[6]], assign = TRUE)

Dirichlet_multinomial_all <- data.frame(cbind(Dirichlet_multinomial_1,
                                              Dirichlet_multinomial_2, Dirichlet_multinomial_3,
                                              Dirichlet_multinomial_4, Dirichlet_multinomial_5,
                                              Dirichlet_multinomial_6))


colnames(Dirichlet_multinomial_all) <- c("DMM_k=1", "DMM_k=2", "DMM_k=3", "DMM_k=4", "DMM_k=5",
                                         "DMM_k=6")

#minimum information criteria
lplc <- sapply(dmn_list, laplace)
aic <- sapply(dmn_list, AIC)
bic <- sapply(dmn_list, BIC)

dmn_list[[which.min(lplc)]] #4, optimal number of metacommunities (Laplace information criterion)
dmn_list[[which.min(aic)]] #1, (Akaike information criterion)
dmn_list[[which.min(bic)]] #1, (Bayesian information criterion)

#plot information criterion
plot(lplc, type = "b", xlab = "Number of Dirichlet Components", ylab = "Model Fit, Laplace")
plot(aic, type = "b", xlab = "Number of Dirichlet Components", ylab = "Model Fit, AIC", lty = 2)
plot(bic, type = "b", xlab = "Number of Dirichlet Components", ylab = "Model Fit, BIC", lty = 3)


fit <- dmn_list
best <- fit[[4]] 

mixturewt(best)
fitted(best, scale = TRUE)
fitted(best)

#want to get influence on the models from the different species
cluster_imp <- fitted(best) #get cluster importance
p1 <- fitted(fit[[1]], scale = TRUE) #comparison with model for one big cluster
p5 <- fitted(best, scale = TRUE)

meandiff <- colSums(abs(p5 - as.vector(p1)))
meandiff

x <- mixture(best)

#plot the community types
plot_list <- list()
for (k in seq(ncol(fitted(best)))) {
  d <- melt(fitted(best))
  colnames(d) <- c("species", "cluster", "value")
  d <- subset(d, cluster == k) %>%
    #arrange OTUs by assingment strength
    arrange(value) %>%
    mutate(species = factor(species, levels = unique(species))) %>%
    #only show the most important drivers
    filter(abs(value) > quantile(abs(value), 0.8))
  
  p <- ggplot(d, aes(x = species, y = value)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(title = paste("Top drivers: community type", k))
  print(p)
}



for (k in seq(ncol(fitted(best)))) {
  d <- melt(fitted(best))
  d$Var1 <- gsub(x = d$Var1, pattern = "\\_", replacement = " ")
  colnames(d) <- c("species", "cluster", "value")
  d <- subset(d, cluster == k) %>%
    #arrange species by assingment strength
    arrange(value) %>%
    mutate(species = factor(species, levels = unique(species))) %>%
    #only show the most important drivers
    filter(abs(value) > quantile(abs(value), 0.8))
  gg_color_hue <- function(n) {
    hues = seq(15, 375, length = n + 1)
    hcl(h = hues, l = 65, c = 100)[1:n]
  }
  cols = brewer.pal(n = 4, name = "Set2")
  #cols = brewer.pal(n = 4, name = "Dark2")
  #cols = brewer.pal(n = 4, name = "Pastel2")
  #cols = c("#8DD3C7", "#BEBADA", "#FB8072", "#80B1D3")
  
  p <- ggplot(d, 
              aes(x = species, y = value)) +
    geom_bar(stat = "identity", fill = cols[k], colour = "black") +
    ylab("Cluster contribution") + xlab("") +
    theme_pubr() +
    theme(axis.text.y = element_text(face = "italic", size = 8)) +
    coord_flip() +
    labs(title = paste("Top drivers: community type", k))
  
  plot_list[[k]] = p
}

plot <- cowplot::plot_grid(plot_list[[1]], plot_list[[2]], plot_list[[3]], 
                           plot_list[[4]],
                           nrow = 2, align = "vh")
plot

#alpa diversity community types
cluster_result <- as.data.frame(Dirichlet_multinomial_4)

cluster_result$shannon <- metadata_single$shannon

stat.test10 <- cluster_result %>%
  dunn_test(shannon ~ Dirichlet_multinomial_4, p.adjust.method = "BH") %>%
  add_xy_position(x = "Dirichlet_multinomial_4")

stat.test10

cluster_result %>%
  kruskal_test(shannon ~ Dirichlet_multinomial_4)

cluster_result %>%
  kruskal_effsize(shannon ~ Dirichlet_multinomial_4, ci = TRUE)


ggplot(cluster_result, aes(x = as.factor(Dirichlet_multinomial_4), y = shannon)) +
  geom_boxplot(width = 0.5
               , fill = cols
               ) +
  geom_point(alpha = 0.6) +
  stat_pvalue_manual(stat.test10, label = "p.adj.signif", tip.length = 0.01, hide.ns = TRUE) +
  xlab("DMM Cluster") + ylab("Shannon diversity") +
  theme_pubr(border = TRUE)


#use previously calculated NMDS to visualize clusters
data.scores$DMMcluster <- cluster_result$Dirichlet_multinomial_4

#cols = c("#8DD3C7", "#BEBADA", "#FB8072" , "#80B1D3"
 #        )

ggplot(data.scores, aes(x = NMDS1, y = NMDS2, color = as.factor(DMMcluster))) +
  geom_point(size = 2.5) +
  stat_ellipse(geom = "polygon", alpha = 0) +
  scale_color_manual(values = cols,
                     name = "DMM cluster") +
  theme_pubr(border = TRUE, legend = "right") 


#calculate relative abundance of species in each DMM cluster
cluster_commu <- RPMM_BE_single_t_filt
cluster_commu$sample <- rownames(cluster_commu)

cluster_abundance <- merge(cluster_commu, cluster_result, by = 0)
rownames(cluster_abundance) <- cluster_abundance$Row.names
cluster_abundance$Row.names <- NULL
cluster_abundance$shannon <- NULL
cluster_abundance$BSI <- NULL
cluster_abundance$sample <- NULL

cluster_abundance$Dirichlet_multinomial_4 <-
  as.factor(cluster_abundance$Dirichlet_multinomial_4)

#mean
cluster_abundance_1 <- apply(cluster_abundance[cluster_abundance$Dirichlet_multinomial_4 == 1,
                                               -ncol(cluster_abundance)], 2, mean)

cluster_abundance_2 <- apply(cluster_abundance[cluster_abundance$Dirichlet_multinomial_4 == 2,
                                               -ncol(cluster_abundance)], 2, mean)

cluster_abundance_3 <- apply(cluster_abundance[cluster_abundance$Dirichlet_multinomial_4 == 3,
                                               -ncol(cluster_abundance)], 2, mean)

cluster_abundance_4 <- apply(cluster_abundance[cluster_abundance$Dirichlet_multinomial_4 == 4,
                                               -ncol(cluster_abundance)], 2, mean)

#merge
cluster_abund_merge <- rbind(cluster_abundance_1, cluster_abundance_2, cluster_abundance_3,
                             cluster_abundance_4)

#calculate and plot relative abundance
cluster_relAbund <- cluster_abund_merge
cluster_relAbund <- as.data.frame(t(cluster_relAbund))
cluster_relAbund <- apply(cluster_abund_merge, 1, function(x) x/sum(x)) 
#colSums(cluster_relAbund) # = 1, 1, 1, 1
cluster_relAbund <- data.frame(cluster_relAbund)

#keep 25 most abundant
cluster_relAbund00 <- cluster_relAbund[c(1:25),]
cluster_relAbund_t <- data.frame(t(cluster_relAbund00))
cluster_relAbund_t$cluster <- rownames(cluster_relAbund_t)

cluster_relAbund_long <- gather(cluster_relAbund_t, key = "species", value = "Abundance",
                                -c(cluster))

cluster_relAbund_long$species <- gsub(x = cluster_relAbund_long$species, 
                                      pattern = "\\_", replacement = " ")

ggplot(cluster_relAbund_long, aes(x = cluster, y = Abundance, fill = species)) +
  geom_bar(position = "fill", stat = "identity", color = "white") +
  scale_y_continuous(expand = c(0.002,0.002),
                     breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c(0, 25, 50, 75, 100)) +
  scale_x_discrete(labels = c("1", "2", "3", "4")) +
  scale_fill_manual(values = c("Gemella haemolysans" = "sandybrown",
                               "Gemella sanguinis" = "sandybrown",
                               #"Haemophilus_influenzae" = "gold",
                               "Haemophilus influenzae" = "gold3",
                               "Lactobacillus gasseri" = "lightgoldenrod1",
                               "Moraxella catarrhalis" = "goldenrod4",
                               "Parvimonas micra" = "khaki3",
                               "Prevotella melaninogenica" = "palegreen3",
                               "Pseudomonas aeruginosa" = "darkgreen",
                               "Rothia dentocariosa" = "darkolivegreen3",
                               "Rothia mucilaginosa" = "darkolivegreen3",
                               "Schaalia odontolytica" = "lightblue2",
                               "Staphylococcus aureus" = "mediumblue",
                               "Streptococcus anginosus" = "lightskyblue",
                               "Streptococcus equinus" = "lightskyblue",
                               "Streptococcus gordonii" = "lightskyblue",
                               "Streptococcus mitis" = "lightskyblue",
                               "Streptococcus oralis" = "lightskyblue",
                               "Streptococcus parasanguinis" = "lightskyblue",
                               "Streptococcus pneumoniae" = "lightskyblue",
                               "Streptococcus pseudopneumoniae" = "lightskyblue",
                               "Streptococcus salivarius" = "lightskyblue",
                               "Streptococcus thermophilus" = "lightskyblue",
                               "Veillonella atypica" = "thistle3",
                               "Veillonella dispar" = "thistle3",
                               "Veillonella parvula" = "thistle3"
                               )) +
  ylab("Mean relative abundance (in %)") + xlab("DMM cluster") +
  #coord_flip() +
  theme_pubr(#legend = "bottom", 
    legend = "right",
    border = TRUE) +
  theme(legend.text = element_text(face = "italic", size = 8)) +
  guides(fill = guide_legend(#nrow = 6
                             ncol = 1
                             ))


#BSI distribution in clusters
cluster_result$BSI <- data.scores$BSI

cluster_result %>%
  kruskal_test(Dirichlet_multinomial_4 ~ BSI)
#p=0.2

cluster_result %>%
  kruskal_test(Dirichlet_multinomial_5 ~ BSI)
#p=0.21


mild_1 <- subset(cluster_result, BSI == "mild" & Dirichlet_multinomial_3 == 1)
mild_2 <- subset(cluster_result, BSI == "mild" & Dirichlet_multinomial_3 == 2)
mild_3 <- subset(cluster_result, BSI == "mild" & Dirichlet_multinomial_3 == 3)
mild_4 <- subset(cluster_result, BSI == "mild" & Dirichlet_multinomial_4 == 4)

moderate_1 <- subset(cluster_result, BSI == "moderate" & Dirichlet_multinomial_3 == 1)
moderate_2 <- subset(cluster_result, BSI == "moderate" & Dirichlet_multinomial_3 == 2)
moderate_3 <- subset(cluster_result, BSI == "moderate" & Dirichlet_multinomial_3 == 3)
moderate_4 <- subset(cluster_result, BSI == "moderate" & Dirichlet_multinomial_4 == 4)

severe_1 <- subset(cluster_result, BSI == "severe" & Dirichlet_multinomial_3 == 1)
severe_2 <- subset(cluster_result, BSI == "severe" & Dirichlet_multinomial_3 == 2)
severe_3 <- subset(cluster_result, BSI == "severe" & Dirichlet_multinomial_3 == 3)
severe_4 <- subset(cluster_result, BSI == "severe" & Dirichlet_multinomial_4 == 4)


ggplot(cluster_result, aes(x = as.factor(Dirichlet_multinomial_4), y = BSI)) +
  geom_bar() 

```

#test metadata in clusters
```{r}
metadata_single$DMM <- cluster_result$Dirichlet_multinomial_4


ggplot(metadata_single, aes(x = as.factor(DMM), y = shannon)) +
  geom_boxplot()

#get frequency of mild/moderate/severe BSI in DMM clusters
table(metadata_single$DMM)
table(metadata_single$BSI)

mild_1 <- subset(metadata_single, BSI == "mild" & DMM == "1")
mild_2 <- subset(metadata_single, BSI == "mild" & DMM == "2")
mild_3 <- subset(metadata_single, BSI == "mild" & DMM == "3")
mild_4 <- subset(metadata_single, BSI == "mild" & DMM == "4")

moderate_1 <- subset(metadata_single, BSI == "moderate" & DMM == "1")
moderate_2 <- subset(metadata_single, BSI == "moderate" & DMM == "2")
moderate_3 <- subset(metadata_single, BSI == "moderate" & DMM == "3")
moderate_4 <- subset(metadata_single, BSI == "moderate" & DMM == "4")

severe_1 <- subset(metadata_single, BSI == "severe" & DMM == "1")
severe_2 <- subset(metadata_single, BSI == "severe" & DMM == "2")
severe_3 <- subset(metadata_single, BSI == "severe" & DMM == "3")
severe_4 <- subset(metadata_single, BSI == "severe" & DMM == "4")

#smoking
table(metadata_single$Smoking_status_.y.n.ex.)

no_1 <- subset(metadata_single, Smoking_status_.y.n.ex. == "0" & DMM == "1")
no_2 <- subset(metadata_single, Smoking_status_.y.n.ex. == "0" & DMM == "2")
no_3 <- subset(metadata_single, Smoking_status_.y.n.ex. == "0" & DMM == "3")
no_4 <- subset(metadata_single, Smoking_status_.y.n.ex. == "0" & DMM == "4")

yes_1 <- subset(metadata_single, Smoking_status_.y.n.ex. == "1" & DMM == "1")
yes_2 <- subset(metadata_single, Smoking_status_.y.n.ex. == "1" & DMM == "2")
yes_3 <- subset(metadata_single, Smoking_status_.y.n.ex. == "1" & DMM == "3")
yes_4 <- subset(metadata_single, Smoking_status_.y.n.ex. == "1" & DMM == "4")

ex_1 <- subset(metadata_single, Smoking_status_.y.n.ex. == "2" & DMM == "1")
ex_2 <- subset(metadata_single, Smoking_status_.y.n.ex. == "2" & DMM == "2")
ex_3 <- subset(metadata_single, Smoking_status_.y.n.ex. == "2" & DMM == "3")
ex_4 <- subset(metadata_single, Smoking_status_.y.n.ex. == "2" & DMM == "4")

#aetiology
table(metadata_single$aetiology)

metadata_single$aet_group <- with(metadata_single,
                                  ifelse(aetiology == "ABPA" | aetiology == "Asthma" |
                                           aetiology == "RheumatoidArthritis" |
                                           aetiology == "InflammatoryBowelDisease", 
                                         "inflammatory",
                                  ifelse(aetiology == "idiopathic", "idiopathic",
                                  ifelse(aetiology == "NTM" | aetiology == "post-infective",
                                         "post-infective",
                                  ifelse(aetiology == "alpha-1AntiTrypsinDeficiency" | 
                                           aetiology == "CFTR-RD" | aetiology == "PCD" |
                                           aetiology == "immunodeficiency", "hereditary",
                                  ifelse(aetiology == "COPD", "COPD", NA))))))

table(metadata_single$aet_group)

COPD_1 <- subset(metadata_single, aet_group == "COPD" & DMM == "1")
COPD_2 <- subset(metadata_single, aet_group == "COPD" & DMM == "2")
COPD_3 <- subset(metadata_single, aet_group == "COPD" & DMM == "3")
COPD_4 <- subset(metadata_single, aet_group == "COPD" & DMM == "4")

hereditary_1 <- subset(metadata_single, aet_group == "hereditary" & DMM == "1")
hereditary_2 <- subset(metadata_single, aet_group == "hereditary" & DMM == "2")
hereditary_3 <- subset(metadata_single, aet_group == "hereditary" & DMM == "3")
hereditary_4 <- subset(metadata_single, aet_group == "hereditary" & DMM == "4")

idiopatic_1 <- subset(metadata_single, aet_group == "idiopathic" & DMM == "1")
idiopatic_2 <- subset(metadata_single, aet_group == "idiopathic" & DMM == "2")
idiopatic_3 <- subset(metadata_single, aet_group == "idiopathic" & DMM == "3")
idiopatic_4 <- subset(metadata_single, aet_group == "idiopathic" & DMM == "4")

inflammatory_1 <- subset(metadata_single, aet_group == "inflammatory" & DMM == "1")
inflammatory_2 <- subset(metadata_single, aet_group == "inflammatory" & DMM == "2")
inflammatory_3 <- subset(metadata_single, aet_group == "inflammatory" & DMM == "3")
inflammatory_4 <- subset(metadata_single, aet_group == "inflammatory" & DMM == "4")

postInfective_1 <- subset(metadata_single, aet_group == "post-infective" & DMM == "1")
postInfective_2 <- subset(metadata_single, aet_group == "post-infective" & DMM == "2")
postInfective_3 <- subset(metadata_single, aet_group == "post-infective" & DMM == "3")
postInfective_4 <- subset(metadata_single, aet_group == "post-infective" & DMM == "4")


#plot distribution of groups in DMM cluster

#DMM_BSI
DMM_BSI <- metadata_single[,c(6,48)]

frame_DMM_BSI <- reshape2::melt(t(apply(table(DMM_BSI$BSI, DMM_BSI$DMM),
                                        1, function(x) x/sum(x))))

ggplot(frame_DMM_BSI, aes(x = Var1, y = as.factor(Var2))) +
  geom_tile(aes(fill = value), color = "black") +
  geom_text(aes(label = scales::percent(value, accuracy = 1))) +
  scale_y_discrete(labels = c("DMM1", "DMM2", "DMM3", "DMM4"),
                   breaks = c("1", "2", "3", "4")) +
  #scale_fill_gradient(high = "#d7301f", low = "#fff7ec") +
  scale_fill_gradient(low = "#FEE0D2", high = "#DE2D26") +
  theme_pubr() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank()) +
  ylab("") + xlab("Bronchiectasis Severity Index") +
  guides(fill = "none")

```


#export dataset for heattree
```{r}
#split dataset into samples belonging to DMM cluster 1, 2, 3 and 4

DMM4 <- as.data.frame(Dirichlet_multinomial_all$`DMM_k=4`)
DMM4$sample <- rownames(Dirichlet_multinomial_all)
rownames(DMM4) <- DMM4$sample
colnames(DMM4) <- c("DMM4", "sample")

RPMM_BE_single_t_filt$DMM <- DMM4$DMM4

RPMM_DMM1 <- subset(RPMM_BE_single_t_filt, RPMM_BE_single_t_filt$DMM == "1")
RPMM_DMM1$DMM <- NULL
RPMM_DMM1_t <- data.frame(t(RPMM_DMM1))

RPMM_DMM2 <- subset(RPMM_BE_single_t_filt, RPMM_BE_single_t_filt$DMM == "2")
RPMM_DMM2$DMM <- NULL
RPMM_DMM2_t <- data.frame(t(RPMM_DMM2))

RPMM_DMM3 <- subset(RPMM_BE_single_t_filt, RPMM_BE_single_t_filt$DMM == "3")
RPMM_DMM3$DMM <- NULL
RPMM_DMM3_t <- data.frame(t(RPMM_DMM3))

RPMM_DMM4 <- subset(RPMM_BE_single_t_filt, RPMM_BE_single_t_filt$DMM == "4")
RPMM_DMM4$DMM <- NULL
RPMM_DMM4_t <- data.frame(t(RPMM_DMM4))

RPMM_BE_single_t_filt$DMM <- NULL

#export tables
write.table(RPMM_DMM1_t, file ="RPMM_DMM1.csv", col.names = TRUE, sep = ";")
write.table(RPMM_DMM2_t, file ="RPMM_DMM2.csv", col.names = TRUE, sep = ";")
write.table(RPMM_DMM3_t, file ="RPMM_DMM3.csv", col.names = TRUE, sep = ";")
write.table(RPMM_DMM4_t, file ="RPMM_DMM4.csv", col.names = TRUE, sep = ";")


#calculate mean relative abundance 
DMM1_rel <- make_relative(as.matrix(RPMM_DMM1))
DMM1_rel <- data.frame(DMM1_rel)

DMM1_rel_t <- data.frame(t(DMM1_rel))
DMM1_rel_t_mean <- as.data.frame(rowMeans(DMM1_rel_t))


DMM2_rel <- make_relative(as.matrix(RPMM_DMM2))
DMM2_rel <- data.frame(DMM2_rel)

DMM2_rel_t <- data.frame(t(DMM2_rel))
DMM2_rel_t_mean <- as.data.frame(rowMeans(DMM2_rel_t))


DMM3_rel <- make_relative(as.matrix(RPMM_DMM3))
DMM3_rel <- data.frame(DMM3_rel)

DMM3_rel_t <- data.frame(t(DMM3_rel))
DMM3_rel_t_mean <- as.data.frame(rowMeans(DMM3_rel_t))


DMM4_rel <- make_relative(as.matrix(RPMM_DMM4))
DMM4_rel <- data.frame(DMM4_rel)

DMM4_rel_t <- data.frame(t(DMM4_rel))
DMM4_rel_t_mean <- as.data.frame(rowMeans(DMM4_rel_t))

sum_DMM_rel <- merge(DMM1_rel_t_mean, DMM2_rel_t_mean, by = 0, all = TRUE)
rownames(sum_DMM_rel) <- sum_DMM_rel$Row.names
sum_DMM_rel$Row.names <- NULL

sum_DMM_rel <- merge(sum_DMM_rel, DMM3_rel_t_mean, by = 0, all = TRUE)
rownames(sum_DMM_rel) <- sum_DMM_rel$Row.names
sum_DMM_rel$Row.names <- NULL

sum_DMM_rel <- merge(sum_DMM_rel, DMM4_rel_t_mean, by = 0, all = TRUE)
rownames(sum_DMM_rel) <- sum_DMM_rel$Row.names
sum_DMM_rel$Row.names <- NULL

colnames(sum_DMM_rel) <- c("DMM1", "DMM2", "DMM3", "DMM4")
sum_DMM_rel <- sum_DMM_rel[order(rowSums(sum_DMM_rel), decreasing = TRUE),]

write.table(sum_DMM_rel, file = "sum_DMM_rel.csv", col.names = TRUE, sep = ";")

#make heattrees for healthy and smokers as well
#get filtered RPMM data

kept_taxa_h <- names(healthy_t)
RPMM_healthy_t_filt <- RPMM_healthy_t[,kept_taxa_h]

kept_taxa_s <- names(smokers_t)
RPMM_healthySmokers_t_filt <- RPMM_healthySmokers_t[,kept_taxa_s]

RPMM_healthy_filt <- data.frame(t(RPMM_healthy_t_filt))
RPMM_healthySmokers_filt <- data.frame(t(RPMM_healthySmokers_t_filt))

write.table(RPMM_healthy_filt, file ="RPMM_healthy.csv", col.names = TRUE, sep = ";")
write.table(RPMM_healthySmokers_filt, file ="RPMM_smokers.csv", col.names = TRUE, sep = ";")

#make combined heattree, rowSums smoker and rowSums healthy
heattree_smokers <- as.data.frame(rowSums(RPMM_healthySmokers_filt))
colnames(heattree_smokers) <- "smoker"

heattree_healthy <- as.data.frame(rowSums(RPMM_healthy_filt))
colnames(heattree_healthy) <- "healthy"

#combine
heattree_healthy_smoker <- merge(heattree_healthy, heattree_smokers, all = TRUE, by = 0)

rownames(heattree_healthy_smoker) <- heattree_healthy_smoker$Row.names
heattree_healthy_smoker$Row.names <- NULL
heattree_healthy_smoker[is.na(heattree_healthy_smoker)] <- 0
heattree_healthy_smoker <- heattree_healthy_smoker[order(rowSums(heattree_healthy_smoker),
                                                         decreasing = TRUE),]

write.table(heattree_healthy_smoker, file ="heattree_healthy_smoker.csv", 
            col.names = TRUE, sep = ";")

#make relAbund rowMeans
smokers_rel_tree <- make_relative(as.matrix(RPMM_healthySmokers_t_filt))
smokers_rel_tree <- data.frame(smokers_rel_tree)

smokers_rel_tree_t <- data.frame(t(smokers_rel_tree))
smokers_rel_tree_t_mean <- as.data.frame(rowMeans(smokers_rel_tree_t))


healthy_rel_tree <- make_relative(as.matrix(RPMM_healthy_t_filt))
healthy_rel_tree <- data.frame(healthy_rel_tree)

healthy_rel_tree_t <- data.frame(t(healthy_rel_tree))
healthy_rel_tree_t_mean <- as.data.frame(rowMeans(healthy_rel_tree_t))


healthy_smokers_tree_relMean <- merge(smokers_rel_tree_t_mean, healthy_rel_tree_t_mean,
                                      by = 0, all = TRUE)
rownames(healthy_smokers_tree_relMean) <- healthy_smokers_tree_relMean$Row.names
healthy_smokers_tree_relMean$Row.names <- NULL
healthy_smokers_tree_relMean[is.na(healthy_smokers_tree_relMean)] <- 0
colnames(healthy_smokers_tree_relMean) <- c("smokers", "healthy")

write.table(healthy_smokers_tree_relMean, 
            file ="heattree_healthy_smoker_relAbundMean.csv", 
            col.names = TRUE, sep = ";")


#combine DMM, healthy and healthy smoker for heattree
heattree_all_rel <- merge(sum_DMM_rel, healthy_smokers_tree_relMean,
                          by = 0, all = TRUE)

rownames(heattree_all_rel) <- heattree_all_rel$Row.names
heattree_all_rel$Row.names <- NULL
heattree_all_rel[is.na(heattree_all_rel)] <- 0
heattree_all_rel <- heattree_all_rel[order(rowSums(heattree_all_rel), decreasing = TRUE),]

write.table(heattree_all_rel,
            file = "heattree_all_relAbundMean.csv",
            col.names = TRUE, sep = ";")

```

#heatmaps based on DMM clustering
```{r}

#calculate age quartiles bronchiectasis
summary(metadata_single$age)
#min 21, 1st Qu 63, Median 68, Mean 66, 3rd Qu. 75, max 91

#1st quartile 21-62 years: 25 patients
#2nd quartile 63-68 years: 26 patients
#3rd quartile 69-74 years: 24 patients
#4th quartile 75-91 years: 26 patients

#make subsets for DMM cluster and age quartiles
metadata_single <- metadata_single[order(rownames(metadata_single)),]
#work with RPMM values

#include cumulative healthy metagenome?
RPMM_healthy_cumulative <- as.data.frame(rowMeans(RPMM_healthy_filt))
colnames(RPMM_healthy_cumulative) <- c("healthy")
RPMM_BE_healthy <- merge(data.frame(t(RPMM_BE_single_t_filt)), RPMM_healthy_cumulative,
                         by = 0, all = TRUE)
rownames(RPMM_BE_healthy) <- RPMM_BE_healthy$Row.names
RPMM_BE_healthy$Row.names <- NULL
RPMM_BE_healthy[is.na(RPMM_BE_healthy)] <- 0
RPMM_BE_healthy <- RPMM_BE_healthy[order(rowSums(RPMM_BE_healthy), decreasing = TRUE),]

heatmap_healthy00 <- as.data.frame(RPMM_BE_healthy[,c(102)])
rownames(heatmap_healthy00) <- rownames(RPMM_BE_healthy)
colnames(heatmap_healthy00) <- c("healthy")
heatmap_healthy01 <- data.frame(t(heatmap_healthy00))
heatmap_healthy01_t <- data.frame(t(heatmap_healthy01))
heatmap_healthy01_t$species <- rownames(heatmap_healthy01_t)

heatmap00 <- as.data.frame(RPMM_BE_healthy[,c(1:101)])
heatmap01 <- data.frame(t(heatmap00))

heatmap01$age <- metadata_single$age
heatmap01$DMM <- cluster_result$Dirichlet_multinomial_4



heatmap_DMM1 <- subset(heatmap01, DMM == "1")
heatmap_DMM2 <- subset(heatmap01, DMM == "2")
heatmap_DMM3 <- subset(heatmap01, DMM == "3")
heatmap_DMM4 <- subset(heatmap01, DMM == "4")

heatmap_DMM1$DMM <- NULL
heatmap_DMM1_1st <- subset(heatmap_DMM1, age < 63)
heatmap_DMM1_2nd <- subset(heatmap_DMM1, age > 62 & age < 69)
heatmap_DMM1_3rd <- subset(heatmap_DMM1, age > 68 & age < 75)
heatmap_DMM1_4th <- subset(heatmap_DMM1, age > 74)

heatmap_DMM1_1st$age <- NULL
heatmap_DMM1_2nd$age <- NULL
heatmap_DMM1_3rd$age <- NULL
heatmap_DMM1_4th$age <- NULL


heatmap_DMM2$DMM <- NULL
heatmap_DMM2_1st <- subset(heatmap_DMM2, age < 63)
heatmap_DMM2_2nd <- subset(heatmap_DMM2, age > 62 & age < 69)
heatmap_DMM2_3rd <- subset(heatmap_DMM2, age > 68 & age < 75)
heatmap_DMM2_4th <- subset(heatmap_DMM2, age > 74)

heatmap_DMM2_1st$age <- NULL
heatmap_DMM2_2nd$age <- NULL
heatmap_DMM2_3rd$age <- NULL
heatmap_DMM2_4th$age <- NULL


heatmap_DMM3$DMM <- NULL
heatmap_DMM3_1st <- subset(heatmap_DMM3, age < 63)
heatmap_DMM3_2nd <- subset(heatmap_DMM3, age > 62 & age < 69)
heatmap_DMM3_3rd <- subset(heatmap_DMM3, age > 68 & age < 75)
heatmap_DMM3_4th <- subset(heatmap_DMM3, age > 74)

heatmap_DMM3_1st$age <- NULL
heatmap_DMM3_2nd$age <- NULL
heatmap_DMM3_3rd$age <- NULL
heatmap_DMM3_4th$age <- NULL


heatmap_DMM4$DMM <- NULL
heatmap_DMM4_1st <- subset(heatmap_DMM4, age < 63)
heatmap_DMM4_2nd <- subset(heatmap_DMM4, age > 62 & age < 69)
heatmap_DMM4_3rd <- subset(heatmap_DMM4, age > 68 & age < 75)
heatmap_DMM4_4th <- subset(heatmap_DMM4, age > 74)

heatmap_DMM4_1st$age <- NULL
heatmap_DMM4_2nd$age <- NULL
heatmap_DMM4_3rd$age <- NULL
heatmap_DMM4_4th$age <- NULL



#randomly select two samples from every cluster age quartile

#rand_DMM1_1st <- heatmap_DMM1_1st[sample(nrow(heatmap_DMM1_1st), 2),] #005,090
rand_DMM1_1st <- heatmap_DMM1_1st[c(1,9),]
rand_DMM1_1st_t <- data.frame(t(rand_DMM1_1st))
rand_DMM1_1st_t$species <- rownames(rand_DMM1_1st_t)
#rand_DMM1_2nd <- heatmap_DMM1_2nd[sample(nrow(heatmap_DMM1_2nd), 2),] #014,029
rand_DMM1_2nd <- heatmap_DMM1_2nd[c(2,3),]
rand_DMM1_2nd_t <- data.frame(t(rand_DMM1_2nd))
rand_DMM1_2nd_t$species <- rownames(rand_DMM1_2nd_t)
#rand_DMM1_3rd <- heatmap_DMM1_3rd[sample(nrow(heatmap_DMM1_3rd), 2),] #013,030
rand_DMM1_3rd <- heatmap_DMM1_3rd[c(1,3),]
rand_DMM1_3rd_t <- data.frame(t(rand_DMM1_3rd))
rand_DMM1_3rd_t$species <- rownames(rand_DMM1_3rd_t)
#rand_DMM1_4th <- heatmap_DMM1_4th[sample(nrow(heatmap_DMM1_4th), 2),] #028,099
rand_DMM1_4th <- heatmap_DMM1_4th[c(3,7),]
rand_DMM1_4th_t <- data.frame(t(rand_DMM1_4th))
rand_DMM1_4th_t$species <- rownames(rand_DMM1_4th_t)

#rand_DMM2_1st <- heatmap_DMM2_1st[sample(nrow(heatmap_DMM2_1st), 2),] #058,087
rand_DMM2_1st <- heatmap_DMM2_1st[c(7,8),]
rand_DMM2_1st_t <- data.frame(t(rand_DMM2_1st))
rand_DMM2_1st_t$species <- rownames(rand_DMM2_1st_t)
#rand_DMM2_2nd <- heatmap_DMM2_2nd[sample(nrow(heatmap_DMM2_2nd), 2),] #092,098
rand_DMM2_2nd <- heatmap_DMM2_2nd[c(8,9),]
rand_DMM2_2nd_t <- data.frame(t(rand_DMM2_2nd))
rand_DMM2_2nd_t$species <- rownames(rand_DMM2_2nd_t)
#rand_DMM2_3rd <- heatmap_DMM2_3rd[sample(nrow(heatmap_DMM2_3rd), 2),] #069,096
rand_DMM2_3rd <- heatmap_DMM2_3rd[c(5,6),]
rand_DMM2_3rd_t <- data.frame(t(rand_DMM2_3rd))
rand_DMM2_3rd_t$species <- rownames(rand_DMM2_3rd_t)
#rand_DMM2_4th <- heatmap_DMM2_4th[sample(nrow(heatmap_DMM2_4th), 2),] #071,095
rand_DMM2_4th <- heatmap_DMM2_4th[c(7,9),]
rand_DMM2_4th_t <- data.frame(t(rand_DMM2_4th))
rand_DMM2_4th_t$species <- rownames(rand_DMM2_4th_t)

#rand_DMM3_1st <- heatmap_DMM3_1st[sample(nrow(heatmap_DMM3_1st), 2),] #019,088
rand_DMM3_1st <- heatmap_DMM3_1st[c(1,2),]
rand_DMM3_1st_t <- data.frame(t(rand_DMM3_1st))
rand_DMM3_1st_t$species <- rownames(rand_DMM3_1st_t)
#rand_DMM3_2nd <- heatmap_DMM3_2nd[sample(nrow(heatmap_DMM3_2nd), 2),] #046,054
rand_DMM3_2nd <- heatmap_DMM3_2nd[c(3,4),]
rand_DMM3_2nd_t <- data.frame(t(rand_DMM3_2nd))
rand_DMM3_2nd_t$species <- rownames(rand_DMM3_2nd_t)
#rand_DMM3_3rd <- heatmap_DMM3_3rd[sample(nrow(heatmap_DMM3_3rd), 2),] #009,070
rand_DMM3_3rd <- heatmap_DMM3_3rd[c(1,2),]
rand_DMM3_3rd_t <- data.frame(t(rand_DMM3_3rd))
rand_DMM3_3rd_t$species <- rownames(rand_DMM3_3rd_t)
#rand_DMM3_4th <- heatmap_DMM3_4th[sample(nrow(heatmap_DMM3_4th), 2),] #085,089
rand_DMM3_4th <- heatmap_DMM3_4th[c(5,6),]
rand_DMM3_4th_t <- data.frame(t(rand_DMM3_4th))
rand_DMM3_4th_t$species <- rownames(rand_DMM3_4th_t)


#rand_DMM4_1st <- heatmap_DMM4_1st[sample(nrow(heatmap_DMM4_1st), 2),] #001,012
rand_DMM4_1st <- heatmap_DMM4_1st[c(1,2),]
rand_DMM4_1st_t <- data.frame(t(rand_DMM4_1st))
rand_DMM4_1st_t$species <- rownames(rand_DMM4_1st_t)
#rand_DMM4_2nd <- heatmap_DMM4_2nd[sample(nrow(heatmap_DMM4_2nd), 2),] #017,102
rand_DMM4_2nd <- heatmap_DMM4_2nd[c(1,3),]
rand_DMM4_2nd_t <- data.frame(t(rand_DMM4_2nd))
rand_DMM4_2nd_t$species <- rownames(rand_DMM4_2nd_t)
#rand_DMM4_3rd <- heatmap_DMM4_3rd[sample(nrow(heatmap_DMM4_3rd), 2),] #068,106
rand_DMM4_3rd <- heatmap_DMM4_3rd[c(1,2),]
rand_DMM4_3rd_t <- data.frame(t(rand_DMM4_3rd))
rand_DMM4_3rd_t$species <- rownames(rand_DMM4_3rd_t)
#rand_DMM4_4th <- heatmap_DMM4_4th[sample(nrow(heatmap_DMM4_4th), 2),] #006,020
rand_DMM4_4th <- heatmap_DMM4_4th[c(1,2),]
rand_DMM4_4th_t <- data.frame(t(rand_DMM4_4th))
rand_DMM4_4th_t$species <- rownames(rand_DMM4_4th_t)

#combine all randomly selected pairs
heatmap_list1 <- list(heatmap_healthy01_t,
                      rand_DMM1_1st_t, rand_DMM1_2nd_t, 
                      rand_DMM1_3rd_t, rand_DMM1_4th_t,
                      rand_DMM2_1st_t, rand_DMM2_2nd_t, 
                      rand_DMM2_3rd_t, rand_DMM2_4th_t,
                      heatmap_healthy01_t,
                      rand_DMM3_1st_t, rand_DMM3_2nd_t, 
                      rand_DMM3_3rd_t, rand_DMM3_4th_t,
                      rand_DMM4_1st_t, rand_DMM4_2nd_t, 
                      rand_DMM4_3rd_t, rand_DMM4_4th_t)
heatmap_combined1 <- Reduce(function(x ,y) merge(x, y, all = TRUE), heatmap_list1)
rownames(heatmap_combined1) <- heatmap_combined1$species
heatmap_combined1$species <- NULL

heatmap_combined1_00 <- heatmap_combined1[order(rowSums(heatmap_combined1), decreasing = TRUE),]

#restrict to 40 most abundant species and keep DMM info
heatmap_combined1_01 <- heatmap_combined1_00[c(1:40),]

#heatmap healthy plus DMM1&2
heatmap_combined_12 <- heatmap_combined1_01[,c(1:17)]

heatmap_combined_12[heatmap_combined_12 == 0] <- NA
rownames(heatmap_combined_12) <- gsub(x = rownames(heatmap_combined_12), 
                                      pattern = "\\_", replacement = " ")

newnames1 <- lapply(rownames(heatmap_combined_12),
                   function(x) bquote(italic(.(x))))

pheatmap(log10(as.matrix((heatmap_combined_12+1))),
         na_col = "grey60",
         border_color = "grey40",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         gaps_col = c(1,3,5,7,9,11,13,15),
         labels_row = as.expression(newnames1),
         fontsize = 9)



#heatmap healthy plus DMM3&4
heatmap_combined_34 <- heatmap_combined1_01[,c(1,18:33)]

heatmap_combined_34[heatmap_combined_34 == 0] <- NA
rownames(heatmap_combined_34) <- gsub(x = rownames(heatmap_combined_34), 
                                      pattern = "\\_", replacement = " ")

newnames2 <- lapply(rownames(heatmap_combined_34),
                   function(x) bquote(italic(.(x))))

pheatmap(log10(as.matrix((heatmap_combined_34+1))),
         na_col = "grey60",
         border_color = "grey40",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         gaps_col = c(1,3,5,7,9,11,13,15),
         labels_row = as.expression(newnames2),
         fontsize = 9)

```



#Supplement: readcount samples and blanks
```{r}

library(scales)

readCounts_BE_healthy_smokers_blanks <- read_excel("readCounts_BE_healthy_smoker_blank.xlsx")

readCounts_BE_healthy_smokers_blanks <- data.frame(readCounts_BE_healthy_smokers_blanks)


#total read counts for WGS analysis

total_readCount <- readCounts_BE_healthy_smokers_blanks[,c(3,6,9,12)]

total_readCount_long <- gather(total_readCount, key = "type", value = "readCount")

#gather(cluster_relAbund_t, key = "species", value = "Abundance", -c(cluster))

total_readCount_long01 <- total_readCount_long[!is.na(total_readCount_long$readCount),]
total_readCount_long01$number <- c(1:228)

ggplot(total_readCount_long01, aes(x = number, y = readCount, color = type)) +
  geom_point() + 
  scale_y_continuous(trans = log10_trans(),
                     breaks = trans_breaks("log10", function(x) 10^x),
                     labels = trans_format("log10", math_format(10^.x))) +
  annotation_logticks(sides = "l") +
  scale_color_manual(values = c("BE_all" = "#F8766D",
                                "blank_all" = "#C77CFF",
                                "healthy_all" = "#7CAE00",
                                "smoker_all" = "#00BFC4"),
                     labels = c("Bronchiectasis", "Blank controls", "Healthy non-smoker",
                                "Healthy smoker"),
                     name = "Sample type") +
  ylab("Total read count") + xlab("Sample") +
  theme_pubr(legend = "right")


#make stacked bar plots for species found in samples and blanks
mean_species_readCount_BE <- read_excel("mean_species_readCount.xlsx", 
    sheet = "BE")
mean_species_readCount_BE <- data.frame(mean_species_readCount_BE)
mean_species_readCount_BE <- mean_species_readCount_BE[order(mean_species_readCount_BE$BE,
                                                             decreasing = TRUE),]


mean_species_readCount_healthy <- read_excel("mean_species_readCount.xlsx", 
    sheet = "healthy")
mean_species_readCount_healthy <- data.frame(mean_species_readCount_healthy)
mean_species_readCount_healthy <- mean_species_readCount_healthy[order(
  mean_species_readCount_healthy$healthy, decreasing = TRUE),]


mean_species_readCount_smoker <- read_excel("mean_species_readCount.xlsx", 
    sheet = "smoker")
mean_species_readCount_smoker <- data.frame(mean_species_readCount_smoker)
mean_species_readCount_smoker <- mean_species_readCount_smoker[order(
  mean_species_readCount_smoker$smoker, decreasing = TRUE),]


mean_species_readCount_blank <- read_excel("mean_species_readCount.xlsx", 
    sheet = "blank")
mean_species_readCount_blank <- data.frame(mean_species_readCount_blank)
mean_species_readCount_blank <- mean_species_readCount_blank[order(
  mean_species_readCount_blank$blank, decreasing = TRUE),]


#take top10 species for each sample type, combine remaining species as "other"
mean_readCount_BE <- mean_species_readCount_BE[c(1:10),]
other_BE <- sum(mean_species_readCount_BE[11:239, "BE"])
mean_readCount_BE[nrow(mean_readCount_BE) + 1,] = c("other", other_BE)

mean_readCount_healthy <- mean_species_readCount_healthy[c(1:10),]
other_healthy <- sum(mean_species_readCount_healthy[11:152, "healthy"])
mean_readCount_healthy[nrow(mean_readCount_healthy) + 1,] = c("other", other_healthy)

mean_readCount_smoker <- mean_species_readCount_smoker[c(1:10),]
other_smoker <- sum(mean_species_readCount_smoker[11:110, "smoker"])
mean_readCount_smoker[nrow(mean_readCount_smoker) + 1,] = c("other", other_smoker)

mean_readCount_blank <- mean_species_readCount_blank[c(1:10),]
#other_blank <- sum(mean_species_readCount_blank[11:110, "blank"])
mean_readCount_blank[nrow(mean_readCount_blank) + 1,] = c("other", "0")

#combine 
mean_readCount_list <- list(mean_readCount_BE, mean_readCount_healthy,
                            mean_readCount_smoker, mean_readCount_blank)

mean_readCount_combined <- Reduce(function(x, y) merge(x, y, all = TRUE), mean_readCount_list)

mean_readCount_combined[is.na(mean_readCount_combined)] <- 0
rownames(mean_readCount_combined) <- mean_readCount_combined$species
mean_readCount_combined$species <- NULL


mean_readCount_combined_t <- data.frame(t(mean_readCount_combined))
mean_readCount_combined_t$type <- rownames(mean_readCount_combined_t)


mean_readCount_combined_long <- gather(mean_readCount_combined_t, key = "species",
                                       value = "readCount",
                                       -c(type))

mean_readCount_combined_long$readCount <- as.numeric(mean_readCount_combined_long$readCount)
mean_readCount_combined_long$species <- gsub(x = mean_readCount_combined_long$species,
                                             pattern = "\\.", replacement = " ")


ggplot(mean_readCount_combined_long, aes(x = type, y = readCount, fill = species)) +
  geom_bar(position = "fill", stat = "identity", color = "white") +
  scale_y_continuous(expand = c(0.002, 0.002),
                     breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c(0, 25, 50, 75, 100)) +
  scale_x_discrete(limits = c("BE", "healthy", "smoker", "blank"),
                   labels = c("Bronchiectasis", "Healthy non-smokers", "Healthy smokers",
                              "Blank controls")) +
  scale_fill_manual(values = c(#"Aeromonas caviae" = "coral4",
                               "Aeromonas caviae" = "bisque4",
                               "Cutibacterium acnes" = "chocolate4",
                               "Gemella sanguinis" = "red3",
                               "Haemophilus influenzae" = "tomato3",
                               "Halomonas campaniensis" = "indianred1",
                               "Kocuria palustris" = "darkorange1",
                               "Micrococcus luteus" = "darkgoldenrod1", 
                               "Moraxella catarrhalis" = "olivedrab1",
                               "Neisseria subflava" = "yellowgreen",
                               "other" = "grey70",
                               "Prevotella jejuni" = "green4",
                               "Prevotella melaninogenica" = "green4",
                               "Pseudomonas aeruginosa" = "seagreen2",
                               "Rothia dentocariosa" = "cadetblue2",
                               "Rothia mucilaginosa" = "cadetblue2",
                               "Schaalia odontolytica" = "turquoise2",
                               "Serratia marcescens" = "royalblue2",
                               "Staphylococcus aureus" = "navy",
                               "Streptococcus equinus" = "purple3",
                               "Streptococcus mitis" = "purple3",
                               "Streptococcus oralis" = "purple3",
                               "Streptococcus parasanguinis" = "purple3",
                               "Streptococcus salivarius" = "purple3",
                               "Veillonella atypica" = "mediumpurple1",
                               "Veillonella dispar" = "mediumpurple1",
                               "Vibrio splendidus" = "plum2"
                               ),
                    name = "Species") +
  ylab("Mean relative abundance (in %)") + xlab("Sample type") +
  theme_pubr(legend = "right") +
  theme(legend.text = element_text(face = "italic", size = 8)) +
  guides(fill = guide_legend(ncol = 1))


ggplot(cluster_relAbund_long, aes(x = cluster, y = Abundance, fill = species)) +
  scale_fill_manual(values = c("Gemella haemolysans" = "sandybrown",
                               "Gemella sanguinis" = "sandybrown",
                               #"Haemophilus_influenzae" = "gold",
                               "Haemophilus influenzae" = "gold3",
                               "Lactobacillus gasseri" = "lightgoldenrod1",
                               "Moraxella catarrhalis" = "goldenrod4",
                               "Parvimonas micra" = "khaki3",
                               "Prevotella melaninogenica" = "palegreen3",
                               "Pseudomonas aeruginosa" = "darkgreen",
                               "Rothia dentocariosa" = "darkolivegreen3",
                               "Rothia mucilaginosa" = "darkolivegreen3",
                               "Schaalia odontolytica" = "lightblue2",
                               "Staphylococcus aureus" = "mediumblue",
                               "Streptococcus anginosus" = "lightskyblue",
                               "Streptococcus equinus" = "lightskyblue",
                               "Streptococcus gordonii" = "lightskyblue",
                               "Streptococcus mitis" = "lightskyblue",
                               "Streptococcus oralis" = "lightskyblue",
                               "Streptococcus parasanguinis" = "lightskyblue",
                               "Streptococcus pneumoniae" = "lightskyblue",
                               "Streptococcus pseudopneumoniae" = "lightskyblue",
                               "Streptococcus salivarius" = "lightskyblue",
                               "Streptococcus thermophilus" = "lightskyblue",
                               "Veillonella atypica" = "thistle3",
                               "Veillonella dispar" = "thistle3",
                               "Veillonella parvula" = "thistle3"
                               )) +
  guides(fill = guide_legend(#nrow = 6
                             ncol = 1
                             ))


```

#comparison wochenende, metaphlan and krakenuniq
```{r}
#load files
metaphlan <- read_excel("merged_abundance_table_bearbeitet.xlsx")
metaphlan <- data.frame(metaphlan)

metaphlan <- metaphlan[-c(1),]
rownames(metaphlan) <- metaphlan$species
metaphlan$species <- NULL
#make new relative table (deleted Homo sapiens)
metaphlan_t <- data.frame(t(metaphlan))
metaphlan_t <- make_relative(as.matrix(metaphlan_t))
metaphlan_t <- data.frame(metaphlan_t)
metaphlan <- data.frame(t(metaphlan_t))
#filtering

#abundance filtering: exclude every species where every column has a threshold below 1%
#abundance > 1% in at least one sample
top_abundant_meta <- metaphlan[!apply(metaphlan<0.01, 1, all, na.rm=TRUE),]
#120 species
#prevalence filtering: species must appear in 10% of all samples
prev_filt_abund = 90

metaphlan_prev <- metaphlan

n_prev = round(length(colnames(metaphlan_prev)) * prev_filt_abund / 100)
metaphlan_prev[metaphlan_prev == 0] <- NA
metaphlan_prev <- metaphlan_prev[rowSums(is.na(metaphlan_prev)) < n_prev, ]
metaphlan_prev[is.na(metaphlan_prev)] <- 0
#63 species

#combine abundance and prevalence filtered datasets
#127 species
metaphlan_filt <- unique(rbind(metaphlan_prev, top_abundant_meta))

metaphlan_filt_t <- data.frame(t(metaphlan_filt))

#make new relative abundance table to again obtain 100%
meta_t <- make_relative(as.matrix(metaphlan_filt_t))
meta_t <- data.frame(meta_t)

meta <- data.frame(t(meta_t))
meta <- meta[order(rowSums(meta), decreasing = TRUE),]

meta_t <- data.frame(t(meta))


wochenende <- read_excel("wochenende_R_filt.xlsx")
wochenende <- data.frame(wochenende)

rownames(wochenende) <- wochenende$species
wochenende$species <- NULL

  
kraken1 <- read_excel("krakenuniq1.xlsx")
kraken1 <- data.frame(kraken1)

rownames(kraken1) <- kraken1$species
kraken1$species <- NULL
kraken1 <- kraken1[-c(1),]

kraken2 <- read_excel("krakenuniq2.xlsx")
kraken2 <- data.frame(kraken2)

rownames(kraken2) <- kraken2$species
kraken2$species <- NULL
kraken2 <- kraken2[-c(1),]

krakenall <- merge(kraken1, kraken2, all = TRUE, by = 0)
rownames(krakenall) <- krakenall$Row.names
krakenall$Row.names <- NULL
krakenall[is.na(krakenall)] <- 0

#make relative table without human
kraken_t <- data.frame(t(krakenall))
kraken_t <- make_relative(as.matrix(kraken_t))
kraken_t <- data.frame(kraken_t)
krakenall <- data.frame(t(kraken_t))

#prevalence and abundance filtering
#abundance filtering: exclude every species where every column has a threshold below 1%
#abundance > 1% in at least one sample
top_abundant_kraken <- krakenall[!apply(krakenall<0.01, 1, all, na.rm=TRUE),]
#145 species
#prevalence filtering: species must appear in 10% of all samples
prev_filt_abund = 90

kraken_prev <- krakenall

n_prev = round(length(colnames(kraken_prev)) * prev_filt_abund / 100)
kraken_prev[kraken_prev == 0] <- NA
kraken_prev <- kraken_prev[rowSums(is.na(kraken_prev)) < n_prev, ]
kraken_prev[is.na(kraken_prev)] <- 0
#226 species

#combine abundance and prevalence filtered datasets
#266 species
kraken_filt <- unique(rbind(kraken_prev, top_abundant_kraken))


#export dataset to remove sp. columns
write.csv(kraken_filt, file = "krakenfilt.csv", row.names = TRUE, col.names = TRUE)

#load table
krakenuniq <- read_delim("krakenfilt_bearbeitet.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

krakenuniq <- data.frame(krakenuniq)
rownames(krakenuniq) <- krakenuniq$species
krakenuniq$species <- NULL

krakenuniq_t <- data.frame(t(krakenuniq))

#make new relative abundance table to again obtain 100%
krakenuniq_t <- make_relative(as.matrix(krakenuniq_t))
krakenuniq_t <- data.frame(krakenuniq_t)

krakenuniq <- data.frame(t(krakenuniq_t))


metadata_all <- read_excel("metadata_comparisonAll.xlsx")
metadata_all <- data.frame(metadata_all)
rownames(metadata_all) <- metadata_all$sample
metadata_all$pair <- as.character(metadata_all$pair)


#combine metaphlan, wochenende and krakenuniq tables
all00 <- merge(krakenuniq, meta, all = TRUE, by = 0)
rownames(all00) <- all00$Row.names
all00$Row.names <- NULL
all00[is.na(all00)] <- 0

all01 <- merge(all00, wochenende, all = TRUE, by = 0)
rownames(all01) <- all01$Row.names
all01$Row.names <- NULL
all01[is.na(all01)] <- 0


all01_t <- data.frame(t(all01))
all01_t <- all01_t[order(rownames(all01_t)),]

#NMDS

set.seed(2)

mdsAll <- metaMDS(all01_t, distance = "bray", k = 3, 
                          autotransform = TRUE, trymax = 100 )
#stress=0.17

stressplot(mdsAll)

data.scoresAll <- as.data.frame(scores(mdsAll)$sites)

metadata_all <-
  metadata_all[order(rownames(metadata_all)),]

data.scoresAll$tool <- metadata_all$tool
data.scoresAll$pair <- metadata_all$pair



envit_metadataAll <- data.scoresAll
envit_metadataAll$NMDS1 <- NULL
envit_metadataAll$NMDS2 <- NULL
envit_metadataAll$NMDS3 <- NULL

envit_metadataAll <- data.frame(cbind(envit_metadataAll, all01_t))

data.envitAll <- envfit(mdsAll, envit_metadataAll, permutation = 999, 
                             na.rm = TRUE)
data.envitAll


ggplot(data.scoresAll, aes(x = NMDS1, y = NMDS2, color = tool)) +
  geom_point(size = 2) +
  #stat_ellipse(geom = "polygon", alpha = 0.1) +
  scale_color_manual(values = c("krakenuniq" = "#7CAE00",
                                "metaphlan" = "#F8766D",
                                "wochenende" = "#00BFC4"),
                     labels = c("KrakenUniq", "MetaPhlAn3", "Wochenende"),
                     name = "Tool") +
  theme_pubr(border = TRUE, legend = "none") 

#add centroids
ggplot(data.scoresAll, aes(x = NMDS1, y = NMDS2, color = tool)) +
  geom_point(size = 2.5, alpha = 0.45) +
  #stat_ellipse(geom = "polygon"
               #, alpha = 0.01
               #) +
  geom_point(aes(x = 0.0939, y = 0.0032), shape = 4, color = "#7CAE00", size = 3) + #krakenuniq
  geom_point(aes(x = -0.0950, y = -0.0071), shape = 4, color = "#F8766D", size = 3) + #metaphla
  geom_point(aes(x = 0.0011, y = 0.0039), shape = 4, color = "#00BFC4", size = 3) + #wochenende
  scale_color_manual(values = c("krakenuniq" = "#7CAE00",
                                "metaphlan" = "#F8766D",
                                "wochenende" = "#00BFC4"),
                     labels = c("KrakenUniq", "MetaPhlAn3", "Wochenende"),
                     name = "Tool") +
  theme_pubr(border = TRUE, legend = "none")

ggplot(data.scoresAll, aes(x = NMDS1, y = NMDS2, color = pair)) +
  geom_point(size = 2) +
  #stat_ellipse(geom = "polygon", alpha = 0.1) +
  theme_pubr(border = TRUE, legend = "none") 



#try to better differentiate points/patient triplets
#try first with subset
subs <- data.scoresAll[c(1:27),]

#test transparent colours
library(colorspace)
## Function for desaturating colors by specified proportion
desat <- function(cols, sat=0.5) {
    X <- diag(c(1, sat, 1)) %*% rgb2hsv(col2rgb(cols))
    hsv(X[1,], X[2,], X[3,])
}

# Utility function for plotting color palettes, 
## (from vignette("hcl-colors"))
pal <- function(col, border = "light gray", ...) {
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
  axes = FALSE, xlab = "", ylab = "", ...)
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}


red100 <- "#E41A1C"
red75 <- desat(red100, 0.75)
red50 <- desat(red100, 0.50)

blue100 <- "#377EB8"
blue75 <- desat(blue100, 0.75)
blue50 <- desat(blue100, 0.50)

green100 <- "#4DAF4A"
green75 <- desat(green100, 0.75)
green50 <- desat(green100, 0.50)

purple100 <- "#984EA3"
purple75 <- desat(purple100, 0.75)
purple50 <- desat(purple100, 0.50)

orange100 <- "#FF7F00"
orange75 <- desat(orange100, 0.75)
orange50 <- desat(orange100, 0.50)

grey100 <- "grey20"
grey75 <- "grey40"
grey50 <- "grey60"


ggplot(subs, aes(x = NMDS1, y = NMDS2, color = pair, shape = pair)) +
  geom_point(size = 5) +
  scale_color_manual(values = c(red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75
                               )) +
  scale_shape_manual(values = c(19, 15, 18, 17, 3, 8,
                                19, 15, 18)) +
  theme_pubr(border = TRUE)

#try for all

ggplot(data.scoresAll, aes(x = NMDS1, y = NMDS2, color = pair, shape = pair)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75, purple75, orange75, grey75,
                                red50, blue50, green50, purple50, orange50, grey50,
                                
                                red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75, purple75, orange75, grey75,
                                red50, blue50, green50, purple50, orange50, grey50,
                                
                                red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75, purple75, orange75, grey75,
                                red50, blue50, green50, purple50, orange50, grey50,
                                
                                red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75, purple75, orange75, grey75,
                                red50, blue50, green50, purple50, orange50, grey50,
                                
                                red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75, purple75, orange75, grey75,
                                red50, blue50, green50, purple50, orange50, grey50,
                                
                                red100, blue100, green100, purple100, 
                                orange100, grey100,
                                red75, blue75, green75, purple75, orange75
                                
                               )) +
  scale_shape_manual(values = c(19, 15, 18, 17, 3, 8,
                                19, 15, 18, 17, 3, 8,
                                19, 15, 18, 17, 3, 8,
                                
                                15, 18, 17, 3, 8, 19,
                                15, 18, 17, 3, 8, 19,
                                15, 18, 17, 3, 8, 19,
                                
                                18, 17, 3, 8, 19, 15,
                                18, 17, 3, 8, 19, 15,
                                18, 17, 3, 8, 19, 15,
                                
                                17, 3, 8, 19, 15, 18,
                                17, 3, 8, 19, 15, 18,
                                17, 3, 8, 19, 15, 18,
                                
                                3, 8, 19, 15, 18, 17,
                                3, 8, 19, 15, 18, 17,
                                3, 8, 19, 15, 18, 17,
                                
                                8, 19, 15, 18, 17, 3,
                                8, 19, 15, 18, 17
                                )) +
  theme_pubr(border = TRUE, legend = "none")

#repeat alpha diversity analysis (mild/moderate/severe BSI)
metadata_all$speciesNumber <- specnumber(all01_t)
metadata_all$shannon <- diversity(all01_t, index = "shannon")


metadata_all %>% 
  kruskal_test(shannon ~ tool)
#p=0.00003
metadata_all %>%
  kruskal_effsize(shannon ~ tool, ci = TRUE) 
  
stat.testAll <- metadata_all %>%
  dunn_test(shannon ~ tool, p.adjust.method = "BH") %>%
  add_xy_position(x = "tool")

stat.testAll

  

ggplot(metadata_all, aes(x = tool, y = shannon)) +
  geom_violin(width = 0.5) +
  geom_point() +
  ylab("Shannon diversity") + xlab("Tool") +
  scale_x_discrete(labels = c("KrakenUniq", "MetaPhlAn3", "Wochenende")) +
  theme_pubr(border = TRUE) +
  stat_pvalue_manual(stat.testAll, label = "p.adj.signif",
                     tip.length = 0.01) +
  #stat_compare_means() +
  geom_pointrange(mapping = aes(x = tool, y = shannon),
                  stat = "summary",
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median,
                  colour = "red", size = 0.3)



kraken_shannon <- subset(metadata_all, tool == "krakenuniq")
kraken_shannon00 <- kraken_shannon$shannon
metaphlan_shannon <- subset(metadata_all, tool == "metaphlan")
metaphlan_shannon00 <-metaphlan_shannon$shannon
wochenende_shannon <- subset(metadata_all, tool == "metaphlan")
wochenende_shannon00 <-wochenende_shannon$shannon

d <- data.frame(kraken_shannon00, metaphlan_shannon00, wochenende_shannon00)
colnames(d) <- c("KrakenUniq", "MetaPhlAn3", "Wochenende")


```

#Maaslin2
```{r}

if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Maaslin2")


library(Maaslin2)

#bronchiectasis vs healthy
input_data <- read.table(file = "BE_healthy_taxonomy.tsv",
                         header = TRUE,
                         sep = "\t",
                         row.names = 1,
                         stringsAsFactors = FALSE)

input_metadata <- read.table(file = "BE_healthy_metadata.tsv",
                             header = TRUE,
                             sep = "\t",
                             stringsAsFactors = FALSE,
                             row.names = 1
                             )

#remove smokers from healthy controls: Sp115, Sp155, Sp281, Sp320, Sp414, Sp416, Sp768, Sp941
smokersToRemove <- c("Sp_115", "Sp_155", "Sp_281", "Sp_320", "Sp_414", "Sp_416", 
                     "Sp_768", "Sp_941")

#bronchiectasis BSI/DMM
input_data <- input_data[!(row.names(input_data) %in% smokersToRemove),]
input_metadata <- input_metadata[!(row.names(input_metadata) %in% smokersToRemove),]

fit.data = Maaslin2(input_data = input_data,
                    input_metadata = input_metadata,
                    min_prevalence = 0,
                    normalization = "NONE",
                    output = "demo_output_healthy_BE",
                    fixed_effects = c("state"))

input_data_BE <- read.table(file = "BE_taxonomy.tsv",
                         header = TRUE,
                         sep = "\t",
                         row.names = 1,
                         stringsAsFactors = FALSE)

input_metadata_BE <- read.table(file = "BE_metadata.tsv",
                             header = TRUE,
                             sep = "\t",
                             stringsAsFactors = FALSE,
                             row.names = 1
                             )

#input_metadata_BE$DMM <- as.character(DMM_resultsForMaaslin)


fit.data_BE = Maaslin2(input_data = input_data_BE,
                    input_metadata = input_metadata_BE,
                    min_prevalence = 0,
                    normalization = "NONE",
                    output = "demo_output_BE_DMM_BSI",
                    fixed_effects = c("BSI", "DMM"),
                    reference = c("BSI,mild", "DMM,1"))

fit.data_BE = Maaslin2(input_data = input_data_BE,
                    input_metadata = input_metadata_BE,
                    min_prevalence = 0,
                    normalization = "NONE",
                    output = "demo_output_BE_BSI_mild",
                    fixed_effects = c("BSI"),
                    reference = c("BSI,mild"))

fit.data_BE = Maaslin2(input_data = input_data_BE,
                    input_metadata = input_metadata_BE,
                    min_prevalence = 0,
                    normalization = "NONE",
                    output = "demo_output_BE_BSI_moderate",
                    fixed_effects = c("BSI"),
                    reference = c("BSI,moderate"))

fit.data_BE = Maaslin2(input_data = input_data_BE,
                    input_metadata = input_metadata_BE,
                    min_prevalence = 0,
                    normalization = "NONE",
                    output = "demo_output_BE_BSI_severe",
                    fixed_effects = c("BSI"),
                    reference = c("BSI,severe"))
```

